---
# Data Analysis Chapter Template
# ================================
# This template captures the pedagogical flow used in Ch 04-06.
# Use it for chapters that teach data manipulation, visualization,
# or statistical analysis using a clinical dataset.
#
# Compared to `programming-chapters.qmd`:
#   - Has an explicit Data Preparation block at the top
#   - Follows a "one tool/concept per section" progression
#   - Includes Interpretation paragraphs after each analysis
#   - Ends with a Common Errors table + Summary journey table
#
# Instructions:
#   1. Copy this file to contents/chapters/XX-chapter-name.qmd
#   2. Replace all [PLACEHOLDERS] with actual content
#   3. Delete sections that don't apply
#   4. Remove this YAML comment block
---

# [Chapter Title]

[1-2 sentence intro: what the reader will learn and why it matters for clinical work.]

[Bridge from previous chapter: "In the previous chapter, we learned X. Now we'll Y — the next step in the workflow."]

```{r}
#| message: false
library(tidyverse)
# [Add chapter-specific packages: gtsummary, ggsci, etc.]
```

::: {.callout-note}
## Our running example: [Scenario name]

[Describe a practical clinical scenario using the dataset. Frame it as a
task the reader might face: "Imagine your colleague asks you to..."
List 3-4 concrete questions the chapter will answer. Each section should
address one of these questions.]

- [Question 1]
- [Question 2]
- [Question 3]
- [Question 4]
:::


## Data Preparation

<!-- PATTERN: Compact prep block
     Ch 04 imports fresh; subsequent chapters replicate the same prep
     in one code chunk with a brief note referencing prior chapters.
     This keeps each chapter self-contained while avoiding re-teaching. -->

[For the first data chapter (e.g., Ch 04), this is the full import section.
For subsequent chapters, condense to a single code chunk:]

We load and prepare the data using tools from [previous chapters]:

```{r}
[dataset]_raw <- read_csv("[path/to/dataset.csv]", show_col_types = FALSE)

[dataset] <- [dataset]_raw |>
  clean_names() |>
  mutate(
    # [Factor conversion, derived variables, etc.]
  )
```

[Brief note: "Everything here — read_csv(), clean_names(), etc. — was taught in Chapter N."]


## [Conceptual Framework / Tool Landscape]

<!-- CONVENTION: Classify to Frame
     Before teaching individual tools, give a bird's-eye overview.
     Use a summary table to show the landscape of what's available.
     This gives the reader a mental map before diving into details.

     Examples from existing chapters:
     - Ch 04: dplyr verbs table (verb | what it does)
     - Ch 05: Grammar of Graphics table (data | aesthetics | geometry)
     - Ch 06: Choosing Summary Statistics table (variable type | distribution | statistic) -->

[Brief intro: "Before diving in, here's the big picture of the tools we'll use:"]

| Category | Tool / Concept | What it does |
|---|---|---|
| **[Category A]** | `tool_a()` | [description] |
| **[Category B]** | `tool_b()` | [description] |
| **[Category C]** | `tool_c()` | [description] |

: [Caption] {.striped}

[Transition: "We'll learn each of these in the sections below, starting with the simplest."]


## [Section 1: Simplest Tool / Concept]

<!-- CONVENTION: One concept per section
     Each section introduces ONE new tool or idea.
     Start with the simplest possible example, then extend.
     The progression within a section:
       1. Basic form (minimal arguments)
       2. Add one option/customization
       3. Add another option
       4. Full polished example -->

### [Basic form]

[Explain what this tool does and when you'd use it.]

```{r}
# [Simplest possible example — minimal arguments]
```

### [First extension]

[Add ONE new option or argument. Explain it before using it.]

```{r}
# [Same example + one new argument]
```

### [Further extension]

```{r}
# [More complete example with multiple options]
```

### Interpretation

<!-- CONVENTION: Interpretation paragraph
     After each major output (table, plot, or summary), write a brief
     clinical interpretation. This models the habit of reading output
     meaningfully rather than just generating it.

     Keep it to 2-3 sentences. Connect the observation to clinical knowledge. -->

[2-3 sentences interpreting the result. What does this tell us clinically?
Is it expected? Does it raise questions for further analysis?]


## [Section 2: Build on Section 1]

<!-- CONVENTION: Incremental Complexity
     Each new section builds on the previous one.
     Assume the reader now knows Section 1's tool.

     Within a section, use the same pattern:
       basic → customized → interpreted

     Examples of incremental progressions:
     - Ch 04: select → filter → mutate (if_else → case_when) → summarise
     - Ch 05: histogram → boxplot → scatter → bar chart
     - Ch 06: single variable → multiple → categorical → grouped → gtsummary -->

### [Simple form with running example]

```{r}
# [Basic usage on the diabetes data]
```

### [Extended form]

<!-- CONVENTION: Explain Syntax Explicitly
     When introducing new syntax, show the general template first,
     then break it down with a bullet list. -->

The syntax is:

```r
# [General syntax template — not evaluated]
```

Let's break this down:

- **`arg1`** --- [what it does]
- **`arg2`** --- [what it does]
- **`arg3`** --- [what it does]

```{r}
# [Applied example with the running dataset]
```

### Interpretation

[2-3 sentences on what the output reveals clinically.]

::: {.callout-caution collapse="true" title="Python Comparison"}
[Python equivalent using pandas/seaborn/etc.]

```python
# Python equivalent
```
:::


## [Section N: Distinguishing Similar Concepts]

<!-- CONVENTION: Clarify confusing distinctions BEFORE they're needed
     When two concepts look similar but behave differently,
     dedicate a callout or mini-section to clarify.

     Examples from existing chapters:
     - Ch 04: if_else() vs ifelse(), factor() vs fct_relevel()
     - Ch 05: fill vs color, + vs |>, geom_col() vs geom_bar()
     - Ch 06: mean+SD vs median+IQR, scale_fill vs scale_color -->

::: {.callout-note}
## `concept_a()` vs `concept_b()`

- **`concept_a()`** --- [when to use it, what it does]
- **`concept_b()`** --- [when to use it, how it differs]

[One-line summary of the key distinction.]
:::


## [Optional: Advanced Shortcut in Collapsed Tip]

<!-- CONVENTION: Defer advanced syntax to optional collapsed callouts
     When there's a more concise/advanced way to do something,
     show the beginner-friendly approach in the main text and
     put the advanced version in a collapsed callout-tip.

     Examples from existing chapters:
     - Ch 06: across() in collapsed tip (column-by-column is primary)
     - Ch 05: ggsci as an optional section at the end
     - Ch 04: select helpers in a tip -->

::: {.callout-tip collapse="true"}
## Shortcut: [Advanced approach name]

[Brief explanation that this is optional and more advanced.]

```{r}
# [Advanced/concise version]
```

[Note that the explicit approach above is perfectly fine for most work.]
:::


## [Optional: Putting It All Together]

<!-- CONVENTION: Pipeline integration
     Near the end, show a multi-step pipeline that combines tools
     from this chapter (and possibly prior chapters) into a single
     realistic workflow. This models how real analysis code looks.

     Examples:
     - Ch 04: filter → mutate → count pipeline
     - Ch 05: filter → mutate → ggplot() pipeline (dplyr into ggplot2)
     - Ch 06: gtsummary with all customizations -->

Let's combine what we've learned into one realistic workflow:

```{r}
# [Multi-step pipeline combining 3+ tools from this chapter]
```

::: {.callout-tip}
## The [pattern name] pattern

[Name the workflow pattern. Example: "transform-then-plot",
"summarise-then-format". This gives readers a mental label
for how to structure their own analyses.]
:::


## Common Errors and Troubleshooting

<!-- CONVENTION: Error table near the end
     Every data analysis chapter includes a troubleshooting table
     with 5-6 common errors specific to this chapter's tools.
     Format: Error Message | Cause | Fix -->

| Error or Symptom | Cause | Fix |
|---|---|---|
| [Error message 1] | [What went wrong] | [How to fix it] |
| [Error message 2] | [What went wrong] | [How to fix it] |
| [Error message 3] | [What went wrong] | [How to fix it] |
| [Error message 4] | [What went wrong] | [How to fix it] |
| [Error message 5] | [What went wrong] | [How to fix it] |

: Common [topic] errors and their fixes {.striped}

::: {.callout-tip}
## The debugging habit ([topic] edition)

[3-step debugging checklist specific to this chapter's tools.
Example for dplyr: "1. Run line by line, 2. Check column names,
3. Check column types with glimpse()"]
:::


## Summary

<!-- CONVENTION: Journey table
     Recap the chapter as a step-by-step journey showing how
     the running example evolved. Each row is a step with:
     - What we did (action)
     - The R tool used
     This mirrors how a real analysis workflow progresses. -->

Here's the journey we took with [the running example]:

| Step | What we did | R tool |
|---|---|---|
| [Step 1] | [Action description] | `tool_1()` |
| [Step 2] | [Action description] | `tool_2()` |
| [Step 3] | [Action description] | `tool_3()` |
| [Step N] | [Action description] | `tool_n()` |

: Chapter N journey --- [from X to Y] {.striped}

[Bridge to next chapter: "In the next chapter, we'll move from X to Y — ..."]


## Exercises {.unnumbered}

<!-- CONVENTION: 4 exercises using the SAME dataset as the chapter.
     Progress from simple (one tool) to complex (multi-step pipeline).

     Pattern:
     1. Simple — apply ONE tool from the chapter
     2. Moderate — combine 2 tools or add interpretation
     3. Applied — build a small analysis addressing a clinical question
     4. Pipeline challenge — combine multiple chapter tools into one workflow

     Each exercise should specify what to compute AND what to interpret. -->

1. **[Simple task name].** [Apply one tool from this chapter. Specify exactly what to compute and from which variables.]

2. **[Moderate task name].** [Combine two tools or concepts. Include a brief interpretation request: "Write 2-3 sentences about what you observe."]

3. **[Applied analysis name].** [A small but complete analysis: prepare data, compute, and interpret. Should feel like a realistic mini-task.]

4. **[Pipeline challenge name].** [Build a multi-step pipeline combining 3+ tools from this chapter. End with: "Read your pipeline aloud — does it make sense?"]
