---
# Programming Chapter Template
# ==============================
# This template implements the pedagogical conventions from CLAUDE.md.
# Use it as a starting point for chapters that teach programming concepts.
#
# Instructions:
#   1. Copy this file to contents/chapters/XX-chapter-name.qmd
#   2. Replace all [PLACEHOLDERS] with actual content
#   3. Delete sections that don't apply
#   4. Remove this YAML comment block
---

# [Chapter Title]

[1-2 sentence intro: what the reader will learn and why it matters.]

[Bridge from the previous chapter: "In the last chapter, we learned X. Now we'll learn Y, which lets us..."]

::: {.callout-note}
## Our running example: [Scenario name]

[Describe the clinical/medical scenario that will thread through this chapter.
Each section should apply its new concept to this same scenario, building
in sophistication (spiral curriculum). Choose a scenario relevant to
healthcare professionals.]
:::


## [Section 1: Simplest Concept]

<!-- CONVENTION: Teach Before Use
     Introduce concepts BEFORE they appear in code.
     If a later section needs operator X, teach it here first. -->

[Explain the concept in plain language, then show the general form.]

[Simple, standalone example first — not the running example yet.]

```{r}
# [Simple example code]
```

[NOW connect to the running example:]

```{r}
# [Running example using this concept]
```

<!-- CONVENTION: Python Comparison (optional, at end of section) -->

::: {.callout-caution collapse="true" title="Python Comparison"}
[Side-by-side R vs Python for this concept.]

```python
# Python equivalent
```
:::


## [Section 2: Build on Section 1]

<!-- CONVENTION: Incremental Complexity
     Introduce the simple form first, then extend.
     Each form gets its own subsection with its own example. -->

### [Simple form]

[Explain and demonstrate the basic version.]

```{r}
# [Basic form with running example]
```

### [Extended form]

[Now introduce the more complex version, building on what the reader just learned.]

```{r}
# [Extended form with running example]
```


## [Section N: New Syntax]

<!-- CONVENTION: Explain Syntax Explicitly
     When introducing new syntax, break it down with bullets. -->

The syntax is:

```r
# [General syntax template]
```

Let's break this down:

- **`part_1`** --- [what it does and why]
- **`part_2`** --- [what it does and why]
- **`part_3`** --- [what it does and why]

[Example applying the syntax to the running example.]


## [Section M: Category of Things]

<!-- CONVENTION: Classify to Frame
     Before details, give a mental model of the landscape. -->

[Brief intro to the category.]

| Category | Description | Example |
|----------|-------------|---------|
| **Type A** | [what it is] | `example_a()` |
| **Type B** | [what it is] | `example_b()` |
| **Type C** | [what it is] | `example_c()` |

[Then teach each category in subsections below.]


## Summary

[Recap the journey: how the running example evolved from start to finish.]

| Step | What we did | Concept learned |
|------|-------------|-----------------|
| 1 | [action] | [concept] |
| 2 | [action] | [concept] |
| ... | ... | ... |

[Bridge to next chapter: "In the next chapter, we'll learn about..."]


## Exercises {.unnumbered}

<!-- Exercises should:
     - Use DIFFERENT medical scenarios than the running example
     - Cover key concepts from the chapter
     - Progress from simple to challenging
     - Solutions go in Appendix C -->

1. **[Exercise name].** [Description using a medical context.]

2. **[Exercise name].** [Description using a medical context.]

3. **[Exercise name].** [Description — can involve writing a function or combining concepts.]

4. **[Exercise name].** [Description — most challenging, combines multiple concepts.]
