# Data Manipulation

In the last chapter, we learned core R syntax and programming concepts. In this chapter, we apply those skills to tabular clinical data: understanding data frames, importing files, manipulating data with `dplyr`, and using tidy data principles.

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(readxl)
library(medicaldata)
```

::: {.callout-note}
## Our running example: From scurvy screening table to diabetes registry cleanup

We will start with a tiny dataset (`scurvy`) to build intuition for data frames and tidy manipulation. Then we will transition to a more realistic workflow using a diabetes registry file (`data/diabetes.csv`, `data/diabetes.xlsx`) for import and analysis tasks.
:::


## What Is a Data Frame/Tibble?

A **vector** stores one type of value in one dimension. A **data frame** stores multiple variables as columns and observations as rows. A **tibble** is the modern tidyverse version of a data frame with cleaner printing and safer defaults.

| Structure | Best for | Example |
|---|---|---|
| Vector | One variable | `c(55, 62, 47)` |
| Data frame | Multiple variables in tabular form | `data.frame(age, sex, glucose)` |
| Tibble | Modern tabular workflow | `tibble(age, sex, glucose)` |

### Simple example first

```{r}
patients_tbl <- tibble(
  patient_id = c("P001", "P002", "P003"),
  age = c(55, 62, 47),
  sex = c("Female", "Male", "Female"),
  glucose_mg_dl = c(145, 132, 178)
)

patients_tbl
```

```{r}
class(patients_tbl)
dim(patients_tbl)
names(patients_tbl)
```

### Explore `scurvy` dataset

Load `scurvy` dataset

```{r}
data("scurvy", package = "medicaldata")
```

```{r}
scurvy |> head()
```


```{r}
class(scurvy)
dim(scurvy)
names(scurvy)
glimpse(scurvy)
```

### Core manipulation syntax

Most tidyverse data manipulation follows this pattern:

```r
dataset |> verb(arguments)
```

- **`dataset`** --- the table you start with
- **`|>`** --- the pipe that sends the left-hand result to the next step
- **`verb(...)`** --- a function that does one clear operation (filter rows, select columns, summarize)

Example:

```{r}
scurvy |>
  select(study_id, treatment)
```

::: {.callout-tip}
## One small verb at a time
Start with one manipulation verb per line. Once each step is clear, combine them into a pipeline.
:::


## Import Data (CSV + Excel Hands-on)

In real projects, data often starts as files. We will import CSV and Excel versions of the same diabetes dataset.

**A) Import CSV with `read_csv()`**

```{r}
diabetes_csv <- read_csv("../../data/diabetes.csv", show_col_types = FALSE)
```

```{r}
head(diabetes_csv)
```

**B) Import Excel with `read_excel()`**

```{r}
diabetes_xlsx <- read_excel("../../data/diabetes.xlsx")
```

```{r}
glimpse(diabetes_xlsx)
```
You can choose to read from a specific sheet or cell range:

```{r}
#| eval: false
# Example: read a specific sheet/range from Excel
read_excel("../../data/diabetes.xlsx", sheet = 1, range = "A1:D20")
```

::: {.callout-warning}
## Common import pitfalls
- **Path error:** `"cannot open the connection"` usually means wrong file path.
- **Missing values encoded oddly:** values like `"NA"`, `"-"`, or blank strings may need explicit cleaning.
- **Type guessing mismatch:** numeric-looking columns may import as text if they include non-numeric tokens.

When troubleshooting, always run `glimpse()` and `summary()` right after import.
:::

::: {.callout-caution collapse="true" title="Python Comparison"}
In Python `pandas`, the equivalent import is:

```python
import pandas as pd

diabetes_csv = pd.read_csv("data/diabetes.csv")
diabetes_xlsx = pd.read_excel("data/diabetes.xlsx")
```

In R, we typically use `readr::read_csv()` and `readxl::read_excel()`.
:::


## Core `dplyr` Verbs

Before this, Let's do automatic data cleaning first with `janitor::clean_names()`, which is a convenience function to clean & standardize column names.

```{r}
diabetes <- medicaldata::diabetes |> janitor::clean_names()
```

Column names before:

```{r}
names(medicaldata::diabetes)
```


Column names after:

```{r}
names(diabetes)
```


What is dplyr? ...


| Category | Verbs | Purpose |
|---|---|---|
| Row operations | `filter()`, `arrange()` | Keep and order patients |
| Column operations | `select()`, `rename()` | Choose and relabel variables |
| Create/transform | `mutate()` | Derive new variables |
| Aggregate | `group_by()`, `summarise()` | Build cohort summaries |

### Column operations: `select()` and `rename()`

```{r}
diabetes |>
  select(glucose_mg_dl, bmi, age, diabetes_5y) |>
  head()
```

```{r}
diabetes |>
  rename(
    age_years = age,
    outcome_5y = diabetes_5y
  ) |>
  head()
```

### Row operations: `filter()` and `arrange()`

```{r}
diabetes |>
  filter(glucose_mg_dl >= 140) |>
  head()
```

```{r}
diabetes |>
  arrange(desc(glucose_mg_dl)) |>
  select(glucose_mg_dl, bmi, age, diabetes_5y) |>
  head()
```

### Create variables: `mutate()`

```{r}
diabetes |>
  mutate(
    high_glucose = glucose_mg_dl >= 140,
    bmi_class = case_when(
      bmi < 18.5 ~ "Underweight",
      bmi < 25 ~ "Normal",
      bmi < 30 ~ "Overweight",
      TRUE ~ "Obesity"
    )
  ) |>
  select(glucose_mg_dl, bmi, high_glucose, bmi_class) |>
  head()
```

### Aggregate with `group_by()` + `summarise()`

```{r}
diabetes |>
  group_by(diabetes_5y) |>
  summarise(
    n = n(),
    mean_glucose = mean(glucose_mg_dl, na.rm = TRUE),
    mean_bmi = mean(bmi, na.rm = TRUE)
  )
```

### Incremental complexity: combine 2-3 verbs

```{r}
diabetes |>
  filter(age >= 30) |>
  mutate(high_glucose = glucose_mg_dl >= 140) |>
  group_by(high_glucose) |>
  summarise(
    n = n(),
    avg_bmi = mean(bmi, na.rm = TRUE)
  )
```

::: {.callout-caution collapse="true" title="Python Comparison"}
A common pandas mental mapping is:

- `filter/select/mutate/group_by + summarise` in R
- `query/loc/assign/groupby().agg()` in pandas

```python
(
    diabetes_csv.query("age >= 30")
    .assign(high_glucose=lambda df: df["glucose_mg-dl"] >= 140)
    .groupby("high_glucose", as_index=False)
    .agg(n=("glucose_mg-dl", "size"), avg_bmi=("bmi", "mean"))
)
```
:::


## Tidy Data Principle (Wide vs Long)

A dataset is **tidy** when:

1. Each variable has its own column.
2. Each observation has its own row.
3. Each value has its own cell.

### Why this matters

Tidy format makes manipulation, summaries, and visualization much easier. Many tidyverse tools are designed around tidy data assumptions.

### Example: from wide to long

```{r}
bp_wide <- tibble(
  patient_id = c("P001", "P002", "P003"),
  bp_day1 = c(142, 130, 150),
  bp_day2 = c(138, 128, 147),
  bp_day3 = c(135, 126, 145)
)

bp_wide
```

```{r}
bp_long <- bp_wide |>
  pivot_longer(
    cols = starts_with("bp_day"),
    names_to = "day",
    values_to = "systolic_bp"
  )

bp_long
```

```{r}
# Optional reverse: long back to wide
bp_long |>
  pivot_wider(
    names_from = day,
    values_from = systolic_bp
  )
```

In the next chapter (visualization), long-format data will help us build cleaner and more flexible plots.


## End-to-End Mini Workflow

Now we combine key steps in one compact pipeline.

```{r}
diabetes |>
  select(glucose_mg_dl, dbp_mm_hg, bmi, age, diabetes_5y) |>
  filter(
    glucose_mg_dl > 0,
    dbp_mm_hg > 0,
    bmi > 0,
    age >= 18
  ) |>
  mutate(
    glucose_group = if_else(glucose_mg_dl >= 140, "High", "Not High"),
    bmi_class = case_when(
      bmi < 25 ~ "Normal",
      bmi < 30 ~ "Overweight",
      TRUE ~ "Obesity"
    )
  ) |>
  group_by(diabetes_5y, glucose_group, bmi_class) |>
  summarise(
    n = n(),
    mean_age = mean(age, na.rm = TRUE),
    .groups = "drop"
  )
```

This is the core workflow pattern you will reuse repeatedly: **import -> inspect -> clean -> transform -> summarize**.


## Summary

| Step | What we did | Concept learned |
|---|---|---|
| 1 | Built and inspected small tables | data frame vs tibble fundamentals |
| 2 | Imported CSV and Excel files | practical data import workflow |
| 3 | Applied core `dplyr` verbs | row/column/transform/aggregate operations |
| 4 | Reshaped wide blood pressure data | tidy data principle (`pivot_longer`) |
| 5 | Built one end-to-end pipeline | reproducible manipulation workflow |

In the next chapter, we will use these cleaned and tidy tables to create clear, publication-ready visualizations with `ggplot2`.


## Exercises {.unnumbered}

1. **ICU vitals filtering.** Create a tibble with `patient_id`, `heart_rate`, `resp_rate`, and `spo2` for 8 patients. Keep only rows with `spo2 < 94` and sort by lowest `spo2` first.

2. **OPD lab import and cleanup.** Import a mock CSV file of outpatient lab values (`hb`, `wbc`, `platelet`, `clinic`). Use `select()` to keep clinically relevant columns and `rename()` to standardize names.

3. **Department-level summary.** Given a table with `department`, `outcome`, and `length_of_stay`, compute counts and mean length of stay grouped by `department` and `outcome`.

4. **Ward-round reshaping task.** Start with a wide table containing `bp_day1`, `bp_day2`, `bp_day3`. Convert it to tidy long format, then compute mean blood pressure by day.

Solutions will be provided in Appendix C.
