# Data Manipulation

In the last chapter, we learned the grammar of R: variables, vectors, functions, and pipes. In this chapter, we switch from *language basics* to *real-world data work* --- the kind of workflow you will use in clinical research and quality improvement projects.

Our focus is simple: take a real diabetes dataset, import it, inspect it, clean it, and summarize it with `dplyr`.

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(readxl)
```

::: {.callout-note}
## Our running example: from file to analysis-ready diabetes registry

Imagine you are preparing a diabetes registry for a quick departmental analysis.

You receive two files from a colleague:

- `data/diabetes.csv`
- `data/diabetes.xlsx`

Your job is to turn those files into a clean table you can trust for summaries and visualization.
:::


## What is a Data Frame?

A **vector** stores one variable.
A **data frame** stores multiple variables as columns, with one patient (or one observation) per row.
A **tibble** is the modern tidyverse version of a data frame.

| Structure | Best for | Example |
|---|---|---|
| Vector | One variable | `c(55, 62, 47)` |
| Data frame | Multiple variables | `data.frame(age, sex, glucose)` |
| Tibble | Modern tabular workflow | `tibble(age, sex, glucose)` |

### Dataframe Creation

You can use either `data.frame()` (base R) or modern alternative [`tibble()`](https://tibble.tidyverse.org/index.html). 

```{r}
patients_tbl <- data.frame(
  patient_id = c("P001", "P002", "P003"),
  age = c(55, 62, 47),
  sex = c("Female", "Male", "Female"),
  glucose_mg_dl = c(145, 132, 178)
)
patients_tbl
```

```{r}
patients_tbl <- tibble(
  patient_id = c("P001", "P002", "P003"),
  age = c(55, 62, 47),
  sex = c("Female", "Male", "Female"),
  glucose_mg_dl = c(145, 132, 178)
)
patients_tbl
```

A quick structural check:

```{r}
class(patients_tbl)
dim(patients_tbl)
names(patients_tbl)
```

When working with any new table, ask three quick questions:

1. What kind of object is this?
2. How many rows and columns does it have?
3. What are the variable names?


## First Contact with a Real Medical Dataset (`scurvy`)

Before touching the diabetes files, let’s inspect one packaged medical dataset to practice orientation.

```{r}
data("scurvy", package = "medicaldata")
scurvy |> head()
```

```{r}
class(scurvy)
dim(scurvy)
names(scurvy)
glimpse(scurvy)
```

`glimpse()` is especially useful because it shows variable names, types, and sample values in one compact view.


## Core Tidyverse Manipulation Pattern

Most table manipulation in this book follows one pattern:

```r
dataset |> verb(arguments)
```

- **`dataset`** --- the table you start with
- **`|>`** --- send output from left to right
- **`verb(...)`** --- one clear operation (`select`, `filter`, `mutate`, `summarise`, etc.)

Example:

```{r}
scurvy |>
  select(study_id, treatment)
```

::: {.callout-tip}
## One verb at a time
When learning, write one verb per line. Once each step is clear, combine them into a pipeline.
:::


## Importing Files: CSV and Excel

In real projects, data usually arrives as files. Let’s import both versions of the diabetes dataset.

### A) Import CSV with `read_csv()`

```{r}
diabetes_raw <- read_csv("../../data/diabetes.csv", show_col_types = FALSE)
```

```{r}
names(diabetes_raw)
nrow(diabetes_raw)
```

```{r}
str(diabetes_raw)
```

```{r}
glimpse(diabetes_raw)
```

```{r}
summary(diabetes_raw)
```

### B) Import Excel with `read_excel()`

```{r}
diabetes_raw <- read_excel("../../data/diabetes.xlsx")
```

```{r}
glimpse(diabetes_raw)
```

You can also read a specific Excel sheet or range:

```{r}
#| eval: false
read_excel("../../data/diabetes.xlsx", sheet = 1)
read_excel("../../data/diabetes.xlsx", range = "A1:D20")
```

::: {.callout-warning}
## Common import pitfalls
- **Wrong path** → file not found
- **Unexpected missing values** (`"NA"`, blanks, custom codes)
- **Wrong variable type** (e.g., numbers imported as text)

After import, check column names & types (`str()`, `glimpse()`), and may be a quick summary (`summary()`) before analysis.
:::

::: {.callout-caution collapse="true" title="Python Comparison"}
In Python/pandas, the equivalent is:

```python
import pandas as pd

diabetes_raw = pd.read_csv("data/diabetes.csv")
diabetes_raw = pd.read_excel("data/diabetes.xlsx")
```
:::


## Prepare Column Names for Comfortable Coding

The raw dataset uses names like `glucose_mg-dl` and `dbp_mm-hg`. These are meaningful, but the hyphen (`-`) is not a best naming convention for dataframe column names.

A common workflow is:

1. Keep source files unchanged (for traceability)
2. Create a working copy with clean names for analysis

```{r}
diabetes <- diabetes_raw |> janitor::clean_names()
```

::: {.callout-tip appearance="default" collapse="true"}
### Tips: `janitor` R package

The `janitor` R package provide useful functionalities for cleaning data in R. The `clean_names()` is one of them. Have a look at <https://sfirke.github.io/janitor/>. 
:::

Column names before and after cleaning:

```{r}
names(diabetes_raw)
names(diabetes)
```

Now we can type cleaner names like `glucose_mg_dl` and `dbp_mm_hg`.


## `dplyr` Verbs: The Clinical Thinking Order

Before syntax, keep this mental model:

- **Which columns matter?** (`select`, `rename`)
- **Which rows matter?** (`filter`, `arrange`)
- **What new variables do I need?** (`mutate`)
- **What cohort-level summary do I need?** (`group_by`, `summarise`)

| Category | Verbs | Purpose |
|---|---|---|
| Column operations | `select()`, `rename()` | Choose and relabel variables |
| Row operations | `filter()`, `arrange()` | Keep and order patients |
| Create/transform | `mutate()` | Derive new variables |
| Aggregate | `group_by()`, `summarise()` | Build cohort summaries |

### 1) Column operations: `select()` and `rename()`

Use `select()` when you want select a **subset** of column. 

```{r}
diabetes |>
  select(glucose_mg_dl, bmi, age, diabetes_5y) |>
  head()
```

Use `rename()` when you want to **rename** a column, to improves readability.

```{r}
diabetes |>
  rename(
    age_years = age,
    outcome_5y = diabetes_5y
  ) |>
  head()
```


### 2) Row operations: `filter()` and `arrange()`

Use `filter()` to keep rows that the condition in `filter(...)` evaluate to `TRUE`. 

```{r}
diabetes |>
  filter(glucose_mg_dl >= 140) |>
  head()
```

Use `arrange()` to **sort** rows by a given column:

```{r}
diabetes |>
  arrange(glucose_mg_dl) |>
  head()
```

To sort from highest first, use `desc()`:

```{r}
diabetes |>
  arrange(desc(glucose_mg_dl)) |>
  head()
```


### 3) Create variables: `mutate()`

```{r}
diabetes |>
  mutate(
    high_glucose = glucose_mg_dl >= 140,
    bmi_class = case_when(
      bmi < 18.5 ~ "Underweight",
      bmi < 25 ~ "Normal",
      bmi < 30 ~ "Overweight",
      TRUE ~ "Obesity"
    )
  ) |>
  select(glucose_mg_dl, bmi, high_glucose, bmi_class) |>
  head()
```

`mutate()` is where raw variables become clinically interpretable variables.

### 4) Aggregate: `group_by()` + `summarise()`

```{r}
diabetes |>
  group_by(diabetes_5y) |>
  summarise(
    n = n(),
    mean_glucose = mean(glucose_mg_dl, na.rm = TRUE),
    mean_bmi = mean(bmi, na.rm = TRUE)
  )
```

This gives a quick cohort-level comparison by 5-year diabetes outcome.

### Build a short pipeline (incremental complexity)

```{r}
diabetes |>
  filter(age >= 30) |>
  mutate(high_glucose = glucose_mg_dl >= 140) |>
  group_by(high_glucose) |>
  summarise(
    n = n(),
    avg_bmi = mean(bmi, na.rm = TRUE)
  )
```

Read it left-to-right:

1. Keep patients age 30+
2. Create a high-glucose flag
3. Compare BMI by glucose group

::: {.callout-caution collapse="true" title="Python Comparison"}
R pipeline idea in pandas:

```python
(
    diabetes_raw.query("age >= 30")
    .assign(high_glucose=lambda df: df["glucose_mg-dl"] >= 140)
    .groupby("high_glucose", as_index=False)
    .agg(n=("glucose_mg-dl", "size"), avg_bmi=("bmi", "mean"))
)
```
:::


## Tidy Data: Wide vs Long

A dataset is **tidy** when:

1. each variable has its own column,
2. each observation has its own row,
3. each value has its own cell.

Why this matters: tidy format makes plotting and grouped analysis much easier.

### Example: blood pressure over multiple days

```{r}
bp_wide <- tibble(
  patient_id = c("P001", "P002", "P003"),
  bp_day1 = c(142, 130, 150),
  bp_day2 = c(138, 128, 147),
  bp_day3 = c(135, 126, 145)
)

bp_wide
```

```{r}
bp_long <- bp_wide |>
  pivot_longer(
    cols = starts_with("bp_day"),
    names_to = "day",
    values_to = "systolic_bp"
  )

bp_long
```

```{r}
# Optional reverse: long -> wide
bp_long |>
  pivot_wider(
    names_from = day,
    values_from = systolic_bp
  )
```

In the next visualization chapter, long format will feel much more natural with `ggplot2`.


## End-to-End Mini Workflow (Realistic Pattern)

Now let’s connect everything into one practical chain.

```{r}
diabetes |>
  select(glucose_mg_dl, dbp_mm_hg, bmi, age, diabetes_5y) |>
  filter(
    glucose_mg_dl > 0,
    dbp_mm_hg > 0,
    bmi > 0,
    age >= 18
  ) |>
  mutate(
    glucose_group = if_else(glucose_mg_dl >= 140, "High", "Not High"),
    bmi_class = case_when(
      bmi < 25 ~ "Normal",
      bmi < 30 ~ "Overweight",
      TRUE ~ "Obesity"
    )
  ) |>
  group_by(diabetes_5y, glucose_group, bmi_class) |>
  summarise(
    n = n(),
    mean_age = mean(age, na.rm = TRUE),
    .groups = "drop"
  )
```

This is a reusable template for many projects:

**import -> inspect -> clean -> transform -> summarize**


## Summary

| Step | What we did | Why it matters |
|---|---|---|
| 1 | Built and inspected small tables | Understand tabular structure before analysis |
| 2 | Imported CSV + Excel diabetes files | Real project entry point |
| 3 | Cleaned column names for coding comfort | Reduce syntax friction and mistakes |
| 4 | Applied core `dplyr` verbs | Core manipulation toolkit |
| 5 | Reshaped wide data to long | Prepare for plotting and grouped analysis |
| 6 | Built one full pipeline | Reproducible workflow you can reuse |

In the next chapter, we will use these cleaned tables to create clear, publication-ready visualizations with `ggplot2`.


## Exercises {.unnumbered}

1. **Quick import check.** Import `../../data/diabetes.csv`, then report:
   - number of rows and columns,
   - variable names,
   - one variable with missing values.

2. **Focused subset for clinic handout.** Create a table with only `age`, `bmi`, `glucose_mg_dl`, and `diabetes_5y`. Keep only rows with `age >= 40` and sort by descending glucose.

3. **Create a risk flag.** Add a new variable `high_risk` defined as:
   - `TRUE` if `glucose_mg_dl >= 140` and `bmi >= 30`
   - `FALSE` otherwise

   Then count how many patients are `high_risk`.

4. **Mini cohort summary.** Build a pipeline that groups by `diabetes_5y` and reports:
   - `n`
   - mean age
   - mean BMI
   - proportion with high glucose (`glucose_mg_dl >= 140`)

Solutions will be provided in Appendix C.
