# Data Frames and Data Manipulation

In the previous chapter, we learned R's grammar --- variables, vectors, functions, and pipes. Now we move from *language basics* to *real data work*: the kind of workflow you'll use daily in clinical research and quality improvement projects.

By the end of this chapter, you'll be able to import a CSV file, inspect it, clean it, transform it, and summarize it --- all with readable, reproducible code.

::: {.callout-note}
## Our running example: Preparing a diabetes registry

Imagine a colleague sends you a file from a diabetes screening program: `diabetes.csv` with 768 patient records. Your job is to turn this raw file into a clean, analysis-ready table.

Along the way, you'll learn how to inspect, import, clean, transform, and summarize tabular data --- the essential skills for any data analysis project.
:::


## What Is a Data Frame?

In @sec-vectors, you learned that a **vector** holds a single variable --- one column of data. A **data frame** is a collection of vectors arranged side-by-side, forming a table with rows and columns. Each column is a variable, each row is an observation (e.g., one patient).

| Structure | What it holds | Example |
|---|---|---|
| Vector | One variable | `c(55, 62, 47)` |
| Data frame | Multiple variables (a table) | A table with `age`, `sex`, `glucose` columns |

### Creating a data frame by hand

You can create a data frame with `data.frame()`:

```{r}
patients <- data.frame(
  patient_id    = c("P001", "P002", "P003"),
  age           = c(55, 62, 47),
  sex           = c("Female", "Male", "Female"),
  glucose_mg_dl = c(145, 132, 178)
)
patients
```

Each argument is a column. Notice that the columns are vectors of the same length --- one value per patient.

### Tibble: the tidyverse data frame

The tidyverse provides `tibble()`, a modern version of `data.frame()`. It prints more neatly and avoids some unexpected behaviors:

```{r}
library(tidyverse)
```

```{r}
patients <- tibble(
  patient_id    = c("P001", "P002", "P003"),
  age           = c(55, 62, 47),
  sex           = c("Female", "Male", "Female"),
  glucose_mg_dl = c(145, 132, 178)
)
patients
```

A tibble shows its dimensions (`3 × 4`) and column types (`<chr>`, `<dbl>`) at the top --- very handy for quick orientation.

::: {.callout-tip}
## The 60-second data orientation

When you encounter any new table, ask three questions:

1. **What kind of object is this?** → `class(data)`
2. **How big is it?** → `dim(data)` (rows × columns)
3. **What are the variable names?** → `names(data)`

```{r}
class(patients)
dim(patients)
names(patients)
```

This quick routine catches most early problems before they become hard-to-trace bugs downstream.
:::

### Exploring structure with `glimpse()`

For a richer overview, `glimpse()` shows every column's name, type, and a few sample values --- all in one compact display:

```{r}
glimpse(patients)
```

You'll use `glimpse()` constantly. It's one of the most useful functions in the tidyverse.


## Warm-Up: Exploring a Packaged Dataset {#sec-warmup-scurvy}

Before touching our diabetes file, let's practice with a small dataset that comes pre-loaded in an R package. The `scurvy` dataset from the `medicaldata` package records a historic clinical trial on scurvy treatments (12 patients, 8 variables).

```{r}
data("scurvy", package = "medicaldata")
```

Apply the 60-second orientation:

```{r}
class(scurvy)
dim(scurvy)
names(scurvy)
```

```{r}
glimpse(scurvy)
```

Notice the column types: `<chr>` for character, `<fct>` for factor. We'll explain factors next --- they're important for clinical data.

::: {.callout-caution collapse="true" title="Python Comparison"}
Python's equivalent of a data frame is the pandas `DataFrame`:

```python
import pandas as pd

# Create a DataFrame
patients = pd.DataFrame({
    "patient_id": ["P001", "P002", "P003"],
    "age": [55, 62, 47],
    "sex": ["Female", "Male", "Female"],
    "glucose_mg_dl": [145, 132, 178]
})

# Orientation
patients.shape    # (3, 4) — like dim()
patients.columns  # column names — like names()
patients.dtypes   # column types — like glimpse()
patients.info()   # combined overview
```
:::


## Factors in Practice {#sec-factors-practice}

In @sec-factors, you learned that a **factor** is R's type for categorical data --- a character variable with a fixed set of **levels**. You also learned to create factors with `factor()` and control level order with the `levels` argument.

Now let's learn a tidyverse tool that makes reordering levels more convenient: `fct_relevel()`.

### Reordering levels with `fct_relevel()`

The `forcats` package (loaded with `library(tidyverse)`) provides `fct_relevel()` to reorder factor levels **after** a factor has been created. This is especially useful when you receive data where the factor order isn't what you need:

```{r}
# Default: alphabetical order
outcome <- factor(c("pos", "neg", "pos", "neg"))
outcome
```

```{r}
# Reorder: put "neg" first (as the reference level)
outcome <- fct_relevel(outcome, "neg", "pos")
outcome
```

`fct_relevel()` takes the factor and the levels in your desired order. You'll use this frequently when preparing data for plots and statistical models, because the first level controls:

- **Plot axis/legend order** --- categories appear in level order
- **Table row order** --- summary tables follow level order
- **Statistical reference category** --- the first level is the baseline in regression

::: {.callout-tip}
## `factor()` vs `fct_relevel()` --- when to use which?

- **`factor(x, levels = ...)`** --- when **creating** a factor from scratch (e.g., from a character vector). You learned this in @sec-factors.
- **`fct_relevel(x, ...)`** --- when **reordering** an existing factor (e.g., a column that's already a factor but in the wrong order).

Both achieve the same result; `fct_relevel()` is simply more convenient for existing factors.
:::


## Importing Data from Files {#sec-import}

In real projects, data arrives as files --- CSV, Excel, or other formats. Let's import our diabetes dataset.

### Reading CSV with `read_csv()`

The `readr` package (part of the tidyverse) provides `read_csv()` for reading CSV files:

```{r}
diabetes_raw <- read_csv("../../data/diabetes.csv", show_col_types = FALSE)
```

Let's break down the syntax:

- **`read_csv("path/to/file.csv")`** --- reads the file at the given path and returns a tibble
- **`show_col_types = FALSE`** --- suppresses the column-type message that `read_csv()` prints by default. Useful for keeping output clean.

::: {.callout-note}
## File paths in a Quarto book

The path `"../../data/diabetes.csv"` means "go up two directories from the chapter file, then into `data/`." In your own R scripts, adjust the path to match where your file lives. Using **RStudio Projects** (as covered in @sec-rstudio-projects) makes paths simpler and more portable.
:::

### Inspecting after import

Always check your data right after import:

```{r}
dim(diabetes_raw)
names(diabetes_raw)
```

```{r}
glimpse(diabetes_raw)
```

We have 768 rows and 9 columns. But notice the column names: `glucose_mg-dl`, `dbp_mm-hg` --- those hyphens will cause problems in R code.

::: {.callout-tip}
## The post-import checklist

Right after reading any file, run these four checks:

1. `dim(data)` --- expected number of rows and columns?
2. `names(data)` --- column names sensible?
3. `glimpse(data)` --- types correct? (numbers as `<dbl>`, text as `<chr>`)
4. `summary(data)` --- any suspicious min/max values or unexpected `NA`s?
:::

### Cleaning column names with `clean_names()`

The `janitor` package provides `clean_names()`, which converts column names to consistent `snake_case` --- replacing hyphens, spaces, and dots with underscores:

```{r}
library(janitor)

diabetes <- diabetes_raw |> clean_names()
names(diabetes)
```

Now `glucose_mg-dl` has become `glucose_mg_dl` --- much easier to type and less error-prone.

::: {.callout-note}
## Why clean names matter

R can handle column names with hyphens (e.g., `` diabetes$`glucose_mg-dl` ``), but you need backticks every time. Clean snake_case names let you write `diabetes$glucose_mg_dl` without any special syntax.
:::

### Reading Excel files (brief) {#sec-import-excel}

For Excel files, the `readxl` package provides `read_excel()`:

```{r}
library(readxl)

diabetes_xlsx <- read_excel("../../data/diabetes.xlsx")
glimpse(diabetes_xlsx)
```

You can also specify a sheet or cell range:

```{r}
#| eval: false
read_excel("data/file.xlsx", sheet = "Sheet2")
read_excel("data/file.xlsx", range = "A1:D20")
```

::: {.callout-warning}
## Common import pitfalls

| Problem | Symptom | Fix |
|---|---|---|
| Wrong file path | `Error: 'file.csv' does not exist` | Check `getwd()`, use RStudio file browser |
| Numbers read as text | Column shows `<chr>` instead of `<dbl>` | Check for text like `"N/A"` in numeric columns |
| Unexpected NAs | More `NA`s than expected | Source file uses blanks or custom codes (`-`, `999`) |
| Encoding issues | Garbled characters | Try `read_csv(..., locale = locale(encoding = "UTF-8"))` |
:::

::: {.callout-caution collapse="true" title="Python Comparison"}
Python's pandas uses similar functions:

```python
import pandas as pd

# CSV
df = pd.read_csv("data/diabetes.csv")

# Excel
df = pd.read_excel("data/diabetes.xlsx", sheet_name="Sheet1")

# Inspect
df.shape
df.columns
df.info()
df.describe()
```
:::


## The dplyr Toolkit: Overview {#sec-dplyr-overview}

Now that our data is imported and cleaned, it's time to **manipulate** it --- selecting columns, filtering rows, creating new variables, and computing summaries.

The `dplyr` package (part of the tidyverse) provides a set of **verbs** --- functions named after what they do. Before diving into each one, here's the big picture:

| Category | Verb | What it does |
|---|---|---|
| **Columns** | `select()` | Choose which columns to keep |
| | `rename()` | Change column names |
| **Rows** | `filter()` | Keep rows that match a condition |
| | `arrange()` | Sort rows by a column |
| **Create** | `mutate()` | Add or modify columns |
| **Aggregate** | `summarise()` | Collapse many rows into a summary |
| | `group_by()` | Split data into groups before summarizing |
| | `count()` | Quick frequency table |

: Core dplyr verbs organized by purpose {.striped}

All dplyr verbs follow the same pattern:

```r
data |> verb(arguments)
```

The first argument is always the data (passed via `|>`), and you refer to column names **without quotes**.


## Selecting and Renaming Columns

### `select()` --- choose columns

Use `select()` to keep only the columns you need:

```{r}
diabetes |>
  select(age, bmi, glucose_mg_dl, diabetes_5y) |>
  head()
```

You can also **exclude** columns with `-`:

```{r}
diabetes |>
  select(-pedigree, -pregnancy_num) |>
  head()
```

::: {.callout-tip}
## Column selection helpers

When you have many columns with similar names, these helpers save typing:

- `starts_with("bp")` --- columns starting with "bp"
- `ends_with("_ml")` --- columns ending with "_ml"
- `contains("glucose")` --- columns containing "glucose"

```{r}
#| eval: false
diabetes |> select(starts_with("insulin"))
```
:::

### `rename()` --- relabel columns

Use `rename()` to give columns clearer names. The syntax is `new_name = old_name`:

```{r}
diabetes |>
  rename(outcome_5y = diabetes_5y) |>
  head(3)
```


## Filtering and Arranging Rows

### `filter()` --- keep rows by condition

`filter()` keeps only the rows where your condition is `TRUE`. This uses the comparison operators you learned in @sec-logical-ops:

```{r}
# Patients with high glucose (>= 140 mg/dL)
diabetes |>
  filter(glucose_mg_dl >= 140) |>
  head()
```

### Combining conditions in `filter()`

Use `&` (AND) or `,` (comma, same as AND) to require multiple conditions. Use `|` (OR) for either-or:

```{r}
# High glucose AND age over 50
diabetes |>
  filter(glucose_mg_dl >= 140, age > 50) |>
  head()
```

```{r}
# Very young OR very old patients
diabetes |>
  filter(age < 25 | age > 65) |>
  head()
```

### `arrange()` --- sort rows

`arrange()` sorts rows by a column in ascending order:

```{r}
diabetes |>
  arrange(glucose_mg_dl) |>
  head()
```

### Sorting in descending order with `desc()`

Wrap a column in `desc()` to sort from highest to lowest:

```{r}
diabetes |>
  arrange(desc(glucose_mg_dl)) |>
  head()
```

`desc()` simply means "descending." You'll use it whenever you want the largest (or latest) values first.


## Creating New Variables with `mutate()` {#sec-mutate}

`mutate()` adds new columns (or modifies existing ones) based on calculations or conditions.

### Simple derived columns

```{r}
diabetes |>
  mutate(age_months = age * 12) |>
  select(age, age_months) |>
  head()
```

The new column `age_months` is computed from the existing `age` column.

### Two-way classification with `if_else()`

Often you need to create a variable with two categories: "yes or no", "high or normal", etc. The `if_else()` function handles this:

```r
if_else(condition, true_value, false_value)
```

Let's break it down:

- **`condition`** --- a logical test applied to each row (e.g., `glucose_mg_dl >= 140`)
- **`true_value`** --- the value to assign when the condition is `TRUE`
- **`false_value`** --- the value to assign when the condition is `FALSE`

```{r}
diabetes |>
  mutate(glucose_flag = if_else(glucose_mg_dl >= 140, "High", "Normal")) |>
  select(glucose_mg_dl, glucose_flag) |>
  head()
```

::: {.callout-tip}
## `if_else()` vs `ifelse()`

In @sec-flow-control, you learned base R's `ifelse()`. The dplyr version `if_else()` (with an underscore) is stricter --- it checks that the true and false values are the same type, catching subtle bugs early. Use `if_else()` inside `mutate()`.
:::

### Multi-way classification with `case_when()` {#sec-case-when}

When you have **more than two categories**, `case_when()` is the tool. Think of it as a vectorized version of `if` / `else if` / `else` from @sec-flow-control:

```r
case_when(
  condition1 ~ value1,
  condition2 ~ value2,
  condition3 ~ value3,
  .default = fallback_value
)
```

Let's break it down:

- Each line is a **condition `~` value** pair (read `~` as "then")
- Conditions are checked **in order** --- the first `TRUE` match wins
- **`.default`** --- the value when no condition matches (like the final `else`)

Here's a clinical example --- classifying BMI:

```{r}
diabetes |>
  mutate(
    bmi_class = case_when(
      bmi < 18.5 ~ "Underweight",
      bmi < 25   ~ "Normal",
      bmi < 30   ~ "Overweight",
      .default   = "Obesity"
    )
  ) |>
  select(bmi, bmi_class) |>
  head(8)
```

A patient with BMI 24.0 doesn't match `< 18.5`, but does match `< 25`, so they get `"Normal"`. The order matters --- just like `if` / `else if` in @sec-flow-control.

::: {.callout-warning}
## The `.default` argument

In older R code and tutorials, you may see `TRUE ~ "Obesity"` instead of `.default = "Obesity"`. Both work, but `.default` is the modern, clearer syntax. We use `.default` throughout this book.
:::

::: {.callout-caution collapse="true" title="Python Comparison"}
Python's pandas offers `np.where()` for two-way (like `if_else()`) and `pd.cut()` or `np.select()` for multi-way (like `case_when()`):

```python
import numpy as np

# Two-way (like if_else)
df["glucose_flag"] = np.where(df["glucose_mg_dl"] >= 140, "High", "Normal")

# Multi-way (like case_when)
conditions = [
    df["bmi"] < 18.5,
    df["bmi"] < 25,
    df["bmi"] < 30,
]
choices = ["Underweight", "Normal", "Overweight"]
df["bmi_class"] = np.select(conditions, choices, default="Obesity")
```
:::


## Summarizing Data {#sec-summarise}

### `summarise()` --- collapse to one row

`summarise()` (or `summarize()` --- both spellings work) computes summary statistics, collapsing the entire table into a single row:

```{r}
diabetes |>
  summarise(
    n_patients   = n(),
    mean_age     = mean(age, na.rm = TRUE),
    mean_glucose = mean(glucose_mg_dl, na.rm = TRUE)
  )
```

Let's break this down:

- **`n()`** --- counts the number of rows (patients)
- **`mean(column, na.rm = TRUE)`** --- computes the mean, removing any `NA` values (recall `na.rm` from @sec-na)
- Each `name = expression` pair becomes a column in the output

### `group_by()` + `summarise()` --- summaries by group

Often you want summaries **per group** rather than for everyone. `group_by()` splits the data, and `summarise()` computes within each group:

```{r}
diabetes |>
  group_by(diabetes_5y) |>
  summarise(
    n            = n(),
    mean_glucose = mean(glucose_mg_dl, na.rm = TRUE),
    mean_bmi     = mean(bmi, na.rm = TRUE)
  )
```

The `pos` group has higher mean glucose --- a good sanity check that the data aligns with clinical expectations.

::: {.callout-note}
## What about `.groups = "drop"`?

After `group_by() |> summarise()`, the result may retain grouping. Adding `.groups = "drop"` at the end of `summarise()` removes it:

```{r}
#| eval: false
diabetes |>
  group_by(diabetes_5y) |>
  summarise(n = n(), .groups = "drop")
```

This is good practice when you plan to pipe the result into further operations. If you see a message about "regrouping," that's R reminding you about this.
:::

### The `count()` shortcut

`count()` is a quick way to get frequency tables --- it combines `group_by()` + `summarise(n = n())` in one step:

```{r}
diabetes |> count(diabetes_5y)
```

You can count combinations of multiple variables:

```{r}
diabetes |>
  mutate(age_group = if_else(age >= 50, "50+", "Under 50")) |>
  count(diabetes_5y, age_group)
```

::: {.callout-caution collapse="true" title="Python Comparison"}
Python's pandas equivalents:

```python
# summarise → agg
df.agg(n_patients=("age", "size"), mean_age=("age", "mean"))

# group_by + summarise → groupby + agg
df.groupby("diabetes_5y").agg(
    n=("age", "size"),
    mean_glucose=("glucose_mg_dl", "mean")
).reset_index()

# count → value_counts
df["diabetes_5y"].value_counts()
```
:::


## Building Pipelines {#sec-building-pipelines}

You've been using `|>` to chain single operations. Now let's build **multi-step pipelines** --- the real power of the tidyverse workflow.

### Start small: two verbs

```{r}
diabetes |>
  filter(age >= 50) |>
  select(age, glucose_mg_dl, diabetes_5y)
```

Read it as: *"Take diabetes, **then** keep patients 50+, **then** select three columns."*

### Add a step: three verbs

```{r}
diabetes |>
  filter(age >= 50) |>
  mutate(glucose_flag = if_else(glucose_mg_dl >= 140, "High", "Normal")) |>
  count(glucose_flag)
```

Each line does one thing. You can read the pipeline top-to-bottom like a recipe.

### Reading a pipeline aloud

A good test: if you can read your pipeline as an English sentence, it's well-written:

> "Take the diabetes data, keep patients aged 50 and older, create a glucose flag (high if >= 140, normal otherwise), then count how many are in each group."

::: {.callout-tip}
## One verb per line

Write each verb on its own line. This makes pipelines easy to read, debug, and modify:

```{r}
#| eval: false
# Good — one verb per line
diabetes |>
  filter(age >= 50) |>
  mutate(glucose_flag = if_else(glucose_mg_dl >= 140, "High", "Normal")) |>
  count(glucose_flag)

# Hard to read — everything on one line
diabetes |> filter(age >= 50) |> mutate(glucose_flag = if_else(glucose_mg_dl >= 140, "High", "Normal")) |> count(glucose_flag)
```
:::

::: {.callout-warning}
## Forgetting `|>` between lines

If you forget a pipe, R treats the next line as a separate (incomplete) command:

```{r}
#| eval: false
# BUG: missing |> after filter()
diabetes |>
  filter(age >= 50)
  select(age, glucose_mg_dl)   # Error!
```

If you see an "unexpected" error, check that every line (except the last) ends with `|>`.
:::


## Tidy Data: Wide vs Long {#sec-tidy-data}

Before we move on, there's one more concept you'll encounter often: **tidy data**. A dataset is **tidy** when:

1. Each **variable** has its own column
2. Each **observation** has its own row
3. Each **value** has its own cell

Why does this matter? Because tidy data makes plotting and grouped analysis much easier. The tools we've learned (`filter()`, `mutate()`, `ggplot2`) all expect tidy format.

### Example: Blood pressure over multiple days

Consider three patients with blood pressure measured on three days:

```{r}
bp_wide <- tibble(
  patient_id = c("P001", "P002", "P003"),
  bp_day1    = c(142, 130, 150),
  bp_day2    = c(138, 128, 147),
  bp_day3    = c(135, 126, 145)
)
bp_wide
```

This is **wide format** --- each day is a separate column. It's readable for humans but awkward for analysis: there's no single "blood pressure" column to plot or summarize.

### `pivot_longer()` --- wide to long

`pivot_longer()` reshapes wide data into tidy (long) format:

```{r}
bp_long <- bp_wide |>
  pivot_longer(
    cols      = starts_with("bp_day"),
    names_to  = "day",
    values_to = "systolic_bp"
  )
bp_long
```

Let's break down the arguments:

- **`cols`** --- which columns to reshape (here, all columns starting with `"bp_day"`)
- **`names_to`** --- name for the new column that will hold the old column names
- **`values_to`** --- name for the new column that will hold the values

Now we have a proper "systolic_bp" variable --- one row per measurement. In @sec-ggplot2, this format will make plotting straightforward.

### `pivot_wider()` --- long to wide (brief)

The reverse operation converts long format back to wide:

```{r}
bp_long |>
  pivot_wider(
    names_from  = day,
    values_from = systolic_bp
  )
```

You'll use `pivot_wider()` less often, but it's good to know it exists.

::: {.callout-note}
## Why tidy format matters

Most R tools --- `ggplot2` for plotting, `dplyr` for summaries, statistical functions --- expect tidy data. When your data isn't in the right shape, `pivot_longer()` and `pivot_wider()` are the tools to fix it.
:::


## Common Errors and Troubleshooting

Here are errors you'll likely encounter when working with data frames and dplyr:

| Error Message | What It Means | Fix |
|---|---|---|
| `object 'colname' not found` | Column name has a typo | Check `names(data)` for the correct spelling |
| `could not find function "select"` | `dplyr` not loaded | Add `library(tidyverse)` at the top |
| `` Error in `filter()`: `` `i` `` must be a logical vector`` | Used `=` instead of `==` in a condition | Use `filter(age == 50)`, not `filter(age = 50)` |
| `'file.csv' does not exist` | Wrong file path | Check `getwd()`, adjust the path |
| `Column 'x' doesn't exist` | Wrong column name after `clean_names()` | Run `names(data)` to see cleaned names |
| `unexpected ')'` | Missing or extra parenthesis | Count your opening and closing parens |

: Common data manipulation errors and their fixes {.striped}

::: {.callout-tip}
## The debugging habit (data edition)

When a dplyr pipeline fails:

1. **Run it line by line** --- add one verb at a time and check the output
2. **Check column names** --- a typo is the most common cause
3. **Check column types** --- `glimpse(data)` shows whether a column is numeric or character
:::


## Summary

Here's the journey we took with the diabetes registry:

| Step | What we did | R tool |
|---|---|---|
| Orient | Explored structure of a table | `class()`, `dim()`, `names()`, `glimpse()` |
| Categorize | Created and ordered factors | `factor()`, `fct_relevel()` |
| Import | Loaded the CSV file | `read_csv()` |
| Clean | Standardized column names | `clean_names()` |
| Select | Chose relevant columns | `select()`, `rename()` |
| Filter | Kept rows by condition | `filter()`, `arrange()`, `desc()` |
| Transform | Created BMI class and glucose flag | `mutate()`, `if_else()`, `case_when()` |
| Summarize | Compared groups | `group_by()`, `summarise()`, `count()` |
| Reshape | Converted wide to long | `pivot_longer()`, `pivot_wider()` |

: Chapter 4 journey --- from raw file to analysis-ready table {.striped}

These data manipulation skills are the foundation for everything that follows. In the next chapter, we'll use these same tools to prepare data for **visualization with ggplot2** --- turning numbers into insights you can see.


## Exercises {.unnumbered}

1. **Quick import check.** Import `diabetes.csv` using `read_csv()`, clean the column names, then report: the number of rows and columns, the column names, and one variable that contains missing values. *(Hint: `summary()` shows `NA` counts.)*

2. **Focused subset for clinic handout.** Starting from the cleaned diabetes table, select only `age`, `bmi`, `glucose_mg_dl`, and `diabetes_5y`. Keep patients aged 40 and older, and sort by descending glucose. Show the first 10 rows.

3. **Blood pressure classification.** Use `mutate()` with `case_when()` to classify `dbp_mm_hg` into:
   - `"Normal"` (< 80)
   - `"Elevated"` (80--89)
   - `"High"` (90 or above)
   - Use `.default` for any remaining values

   Then use `count()` to see how many patients fall in each category.

4. **Cohort comparison pipeline.** Build a pipeline that:
   a. Filters to patients aged 30 and above
   b. Creates a `glucose_flag` using `if_else()` (High if `glucose_mg_dl >= 140`, Normal otherwise)
   c. Groups by `diabetes_5y` and `glucose_flag`
   d. Summarizes the count and mean BMI per group

   Read your pipeline aloud --- does it make sense as an English sentence?
