# Data Visualization with ggplot2

In the previous chapter, we learned how to import, inspect, and transform data. In this chapter, we move to the next core skill: **visualization**.

For many beginners, `ggplot2` feels difficult at first because of its layered syntax. That is normal. We will move step by step and explain every new concept before using it.

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(janitor)
library(ggsci)
library(palmerpenguins)
```

::: {.callout-note}
## Our running example: visualizing a diabetes registry

Throughout this chapter, we will use `data/diabetes.csv` as our main dataset.

Our guiding questions are practical:

- How are glucose values distributed?
- How do distributions differ by diabetes outcome?
- How does BMI relate to glucose?
- How do we present this clearly in a professional setting?
:::


## Why Visualization Matters in Clinical Work

A table is precise, but a figure is often faster to interpret. Good visualizations help us:

1. detect patterns and outliers,
2. compare groups,
3. communicate clearly with teams,
4. prepare publication- or presentation-ready outputs.

In clinical analytics, clear figures reduce misunderstanding and improve decision-making.


## Before Plotting: Prepare the Data

```{r}
diabetes_raw <- read_csv("../../data/diabetes.csv", show_col_types = FALSE)

diabetes <- diabetes_raw |>
  clean_names() |>
  mutate(
    diabetes_5y = fct_relevel(diabetes_5y, "neg", "pos"),
    bmi_class = case_when(
      bmi < 18.5 ~ "Underweight",
      bmi < 25 ~ "Normal",
      bmi < 30 ~ "Overweight",
      TRUE ~ "Obesity"
    ) |> factor(levels = c("Underweight", "Normal", "Overweight", "Obesity"))
  )

glimpse(diabetes)
```

### New helper functions used above

We introduced three functions that are important for plotting workflows:

- **`clean_names()`** (`janitor`)  
  standardizes column names to a coding-friendly style.
- **`fct_relevel()`** (`forcats`)  
  sets factor order explicitly (so plot order is controlled).
- **`drop_na()`** (`tidyr`, used later)  
  removes rows with missing values in selected columns.

::: {.callout-tip}
## Practical pre-plot checklist

Before making any plot, check:

1. `names(data)`
2. `glimpse(data)`
3. `summary(data)`

This catches many errors early.
:::


## The ggplot2 Mental Model (Teach Before Use)

Most plots follow this structure:

```r
ggplot(data = my_data, aes(x = ..., y = ...)) +
  geom_*()
```

Let’s break this down:

- **`ggplot(data = ...)`** --- choose the dataset.
- **`aes(...)`** --- map variables to visual properties.
  - `x` and `y` are most common.
  - `color` and `fill` are also common.
- **`geom_*()`** --- choose how to draw marks (bars, points, boxes, etc.).
- **`+`** --- add layers to the plot.

Think: **data + mapping + geometry + layers**.

### Why `+` appears at the end of lines

In `ggplot2`, each `+` means "add the next layer." Writing one layer per line makes code easier to read.

```r
ggplot(...) +
  geom_...() +
  labs(...) +
  theme_...()
```


## First Plot: Histogram of Glucose

A histogram is a strong first choice for a continuous variable.

```{r}
ggplot(diabetes, aes(x = glucose_mg_dl)) +
  geom_histogram(binwidth = 10, fill = "#2C7FB8", color = "white", alpha = 0.9) +
  geom_vline(xintercept = 140, linetype = "dashed", linewidth = 1, color = "#D95F02") +
  annotate("text", x = 142, y = 85, label = "140 mg/dL", hjust = 0, color = "#D95F02") +
  labs(
    title = "Distribution of Plasma Glucose",
    subtitle = "Diabetes dataset (Pima)",
    x = "Glucose (mg/dL)",
    y = "Number of patients"
  ) +
  theme_minimal(base_size = 12)
```

### New syntax in this plot

- **`geom_histogram(binwidth = 10)`**: create histogram bars with width 10 units.
- **`geom_vline(xintercept = 140)`**: draw vertical reference line at x = 140.
- **`annotate("text", ...)`**: place custom text on plot.
- **`labs(...)`**: set title, subtitle, axis labels.
- **`theme_minimal()`**: apply a clean visual theme.

Interpretation: this plot shows distribution shape and spread. The dashed 140 mg/dL line is a useful clinical reference.


## Group Comparison: Boxplot + Jitter

When comparing a continuous variable across groups, boxplots are often the best starting point.

```{r}
diabetes_box <- diabetes |>
  drop_na(glucose_mg_dl, diabetes_5y)

ggplot(diabetes_box, aes(x = diabetes_5y, y = glucose_mg_dl, fill = diabetes_5y)) +
  geom_boxplot(alpha = 0.75, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.25, size = 1, color = "#333333") +
  scale_fill_manual(values = c("neg" = "#4DAF4A", "pos" = "#E41A1C")) +
  labs(
    title = "Glucose by 5-Year Diabetes Outcome",
    x = "Outcome at 5 years",
    y = "Glucose (mg/dL)",
    fill = "Outcome"
  ) +
  theme_minimal(base_size = 12)
```

### New syntax in this plot

- **`drop_na(col1, col2)`**: keep only rows where selected columns are not missing.
- **`geom_boxplot()`**: show median, quartiles, and spread.
- **`geom_jitter(width = ...)`**: add points with slight horizontal noise to reduce overlap.
- **`scale_fill_manual(...)`**: choose custom colors for groups.

Interpretation: the `pos` group tends to have higher glucose values than the `neg` group.


## Relationship Plot: Scatter + Trend Line

Use a scatter plot to examine two numeric variables together.

```{r}
diabetes_scatter <- diabetes |>
  drop_na(bmi, glucose_mg_dl, diabetes_5y)

ggplot(diabetes_scatter, aes(x = bmi, y = glucose_mg_dl, color = diabetes_5y)) +
  geom_point(alpha = 0.65, size = 2) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1) +
  scale_color_manual(values = c("neg" = "#1B9E77", "pos" = "#D95F02")) +
  labs(
    title = "Relationship Between BMI and Glucose",
    subtitle = "Colored by 5-year diabetes outcome",
    x = "Body mass index (kg/m²)",
    y = "Glucose (mg/dL)",
    color = "Outcome"
  ) +
  theme_minimal(base_size = 12)
```

### New syntax in this plot

- **`geom_point()`**: draw one point per observation.
- **`geom_smooth(method = "lm", se = TRUE)`**: add a **linear** trend line with confidence band.
- **`se = TRUE`**: show the confidence band around the fitted line.
- **`scale_color_manual(...)`**: set manual colors for `color` aesthetic.

Notable option to explore:

- If you write `geom_smooth()` **without** `method = "lm"`, ggplot2 will use its default smoothing method (typically a non-linear smoother for smaller datasets). This is useful for exploratory visualization when relationships may not be linear.

Interpretation: the trend appears positive (higher BMI associated with higher glucose), but this is descriptive, not causal.


## Categorical Comparison: Bar Chart

Now we compare counts across category combinations.

```{r}
bmi_counts <- diabetes |>
  count(diabetes_5y, bmi_class)

ggplot(bmi_counts, aes(x = bmi_class, y = n, fill = diabetes_5y)) +
  geom_col(position = "dodge") +
  labs(
    title = "BMI Class Distribution by 5-Year Diabetes Outcome",
    x = "BMI class",
    y = "Number of patients",
    fill = "Outcome"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

### New syntax in this plot

- **`count(var1, var2)`**: quick grouped counts.
- **`geom_col()`**: bar chart using precomputed y values.
- **`position = "dodge"`**: place bars side-by-side for comparison.

Notable options to explore:

- **`position = "stack"`** (default): stack groups in one bar.
- **`position = "fill"`**: show proportions (each bar scaled to 100%).
- **`position = "identity"`**: draw bars on top of each other (usually with transparency).

This plot communicates subgroup composition more clearly than a printed table.


## Faceting: One Figure, Multiple Panels

Faceting splits one plotting template into several panels by subgroup.

```{r}
diabetes_facet <- diabetes |>
  drop_na(glucose_mg_dl, diabetes_5y)

ggplot(diabetes_facet, aes(x = glucose_mg_dl)) +
  geom_histogram(binwidth = 10, fill = "#3B528B", color = "white") +
  facet_wrap(~ diabetes_5y, nrow = 1) +
  labs(
    title = "Glucose Distribution Stratified by Outcome",
    x = "Glucose (mg/dL)",
    y = "Number of patients"
  ) +
  theme_minimal(base_size = 12)
```

### New syntax in this plot

- **`facet_wrap(~ group_var)`**: create separate panels by a grouping variable.

Faceting reduces clutter and makes within-group patterns easier to inspect.


## Professional Styling Checklist

Before sharing a figure in slides or reports, verify:

- title is informative,
- axes include units,
- legend labels are clear,
- colors are interpretable and consistent,
- theme is readable.

Good analysis with unclear visuals is still weak communication.


## A "Wow" Figure with ggsci (Palmer Penguins)

Near the end of the chapter, we now add a style-focused example. The goal is to demonstrate what a polished, scientific-looking figure can look like.

```{r}
penguins_clean <- penguins |>
  drop_na(bill_length_mm, flipper_length_mm, species)

ggplot(penguins_clean, aes(x = bill_length_mm, y = flipper_length_mm, color = species)) +
  geom_point(alpha = 0.9, size = 2.4) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.9) +
  scale_color_lancet() +
  labs(
    title = "Palmer Penguins: Bill Length vs Flipper Length",
    subtitle = "Styled with ggsci Lancet palette for publication-like aesthetics",
    x = "Bill length (mm)",
    y = "Flipper length (mm)",
    color = "Species",
    caption = "Data: palmerpenguins"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(size = 11, color = "#555555"),
    panel.grid.minor = element_blank(),
    legend.position = "top"
  )
```

### New syntax in this plot

- **`scale_color_lancet()`** (`ggsci`): apply Lancet-inspired color palette.
- **`geom_smooth(..., se = FALSE)`**: show the fitted line without confidence band (cleaner presentation style).
- **`theme(...)`**: fine-tune specific visual elements (title font, grid lines, legend position).

This example is intentionally style-forward: fundamentals first, then refinement.


## End-to-End Mini Workflow

Now combine transform + summarize + visualize in one pipeline:

```{r}
diabetes |>
  filter(glucose_mg_dl > 0, bmi > 0) |>
  mutate(glucose_group = if_else(glucose_mg_dl >= 140, "High", "Not High")) |>
  group_by(diabetes_5y, glucose_group) |>
  summarise(n = n(), .groups = "drop") |>
  ggplot(aes(x = diabetes_5y, y = n, fill = glucose_group)) +
  geom_col(position = "dodge") +
  labs(
    title = "Count of High vs Non-High Glucose by 5-Year Outcome",
    x = "Outcome at 5 years",
    y = "Number of patients",
    fill = "Glucose group"
  ) +
  theme_minimal(base_size = 12)
```

This pattern is practical for real reporting:

**filter/transform -> summarise -> visualize**


## Summary

| Step | What we learned | Why it matters |
|---|---|---|
| 1 | ggplot mental model | Reduces syntax confusion |
| 2 | Histogram and boxplot | Understand distributions and group differences |
| 3 | Scatter + trend line | Explore variable relationships |
| 4 | Bar chart and faceting | Compare categories and subgroups clearly |
| 5 | Styling with ggsci | Improve professional presentation quality |

In the next chapter, we will move from visual summaries to formal **descriptive statistics**.


## Exercises {.unnumbered}

1. **Histogram practice.** Create a histogram of `glucose_mg_dl` with bin width 5 and add a vertical line at 140 mg/dL.

2. **Boxplot interpretation.** Create a boxplot of `bmi` by `diabetes_5y` with jitter points. Write 2–3 sentences describing what you observe.

3. **Scatter with trend.** Plot `age` vs `glucose_mg_dl`, color by `diabetes_5y`, and add a linear trend line.

4. **Professional styling challenge.** Recreate one diabetes plot using a `ggsci` palette (e.g., Lancet or NPG), with polished title, labels, and caption for a conference slide.

Solutions will be provided in Appendix C.
