# Data Visualization with ggplot2 {#sec-ggplot2}

In the previous chapter, we learned to import, clean, and transform data. Now it's time to **see** it. A well-made figure often reveals patterns that tables alone cannot --- outliers, skewed distributions, group differences, and relationships between variables.

For many beginners, `ggplot2` feels unusual at first because of its layered syntax. That's normal. We'll build up one layer at a time, explaining each new piece before using it.

```{r}
#| message: false
library(tidyverse)
library(janitor)
```

::: {.callout-note}
## Our running example: Visualizing a diabetes cohort

Throughout this chapter, we'll use the diabetes dataset to answer practical questions:

- How are glucose values distributed in this cohort?
- Do distributions differ by diabetes outcome?
- How does BMI relate to glucose?
- How can we present findings clearly for a meeting or publication?

Each plot will answer one question, and each section introduces one new ggplot2 concept.
:::


## Why Visualization Matters in Clinical Work

A table of 768 rows is precise, but a figure is faster to interpret. Good visualizations help us:

1. Detect **patterns and outliers** that numbers alone may hide
2. **Compare groups** at a glance
3. **Communicate clearly** with clinical teams and stakeholders
4. Prepare **publication-** or **presentation-ready** outputs

In clinical analytics, a clear figure can prevent misunderstandings that a dense table might cause.


## Data Preparation

We load and prepare the diabetes data using tools from @sec-import and @sec-mutate. This block applies the same steps you learned in Chapter 4:

```{r}
diabetes_raw <- read_csv("../../data/diabetes.csv", show_col_types = FALSE)

diabetes <- diabetes_raw |>
  clean_names() |>
  mutate(
    diabetes_5y = fct_relevel(diabetes_5y, "neg", "pos"),
    bmi_class = case_when(
      bmi < 18.5 ~ "Underweight",
      bmi < 25   ~ "Normal",
      bmi < 30   ~ "Overweight",
      .default   = "Obesity"
    ) |> factor(levels = c("Underweight", "Normal", "Overweight", "Obesity"))
  )
```

Everything here --- `read_csv()`, `clean_names()`, `fct_relevel()`, `case_when()`, `factor()` --- was taught in the previous chapter.


## The Grammar of Graphics {#sec-grammar-of-graphics}

`ggplot2` is built on a simple idea: every plot is made of **layers**, and each layer has three components:

| Component | What it does | Example |
|---|---|---|
| **Data** | The dataset to plot | `diabetes` |
| **Aesthetics** (`aes`) | Map variables to visual properties | `x = age`, `y = glucose_mg_dl`, `color = diabetes_5y` |
| **Geometry** (`geom`) | How to draw the data | Points, bars, boxes, lines |

: The three ingredients of a ggplot2 plot {.striped}

### The minimal plot template

Every ggplot2 plot starts with this pattern:

```r
ggplot(data, aes(x = ..., y = ...)) +
  geom_*()
```

Let's break it down:

- **`ggplot(data, aes(...))`** --- creates the plot canvas. `data` is your tibble, `aes()` maps columns to visual properties.
- **`aes(x = ..., y = ...)`** --- the **aesthetic mapping**. Think of it as "which variable goes where." Common mappings: `x`, `y`, `color`, `fill`, `size`.
- **`+ geom_*()`** --- adds a **geometry layer** that draws the data (histogram bars, scatter points, box plots, etc.)
- **`+`** --- the "add a layer" operator. Each `+` adds something to the plot.

### Why `+` instead of `|>`?

In data manipulation, we use `|>` to chain operations. In ggplot2, we use `+` to add layers. They look similar but mean different things:

- **`|>`** = "**then** do this to the data"
- **`+`** = "**also** draw this on the plot"

::: {.callout-warning}
## Don't mix `+` and `|>`

A common beginner mistake is using `|>` inside ggplot2 code. Remember: `|>` is for data pipelines, `+` is for plot layers. If you see an error like *"object '+' not found"* or *"did you accidentally use `|>` instead of `+`"*, check your operators.
:::

::: {.callout-caution collapse="true" title="Python Comparison"}
Python's `matplotlib` and `seaborn` take a different approach --- instead of layering with `+`, you call separate functions:

```python
import matplotlib.pyplot as plt
import seaborn as sns

# seaborn (high-level, similar philosophy to ggplot2)
sns.histplot(data=diabetes, x="glucose_mg_dl", bins=20)
plt.title("Distribution of Glucose")
plt.show()
```

R's `ggplot2` and Python's `seaborn` share the same intellectual ancestor --- Leland Wilkinson's *Grammar of Graphics*.
:::


## Your First Plot: Histogram

A histogram shows the **distribution** of a single continuous variable. Let's start with the simplest possible version and build up.

### A basic histogram

```{r}
ggplot(diabetes, aes(x = glucose_mg_dl)) +
  geom_histogram()
```

That's it --- three lines. R picked a default bin width and drew the bars. But we can do much better.

### Adjusting bin width

The default bin count (30) may not suit your data. Use `binwidth` to control the width of each bar:

```{r}
ggplot(diabetes, aes(x = glucose_mg_dl)) +
  geom_histogram(binwidth = 10)
```

Each bar now spans 10 mg/dL. Experiment with different values --- smaller bins show more detail, larger bins show the overall shape.

### Adding color and labels

::: {.callout-note}
## `fill` vs `color` in ggplot2

These two aesthetics are easy to confuse:

- **`fill`** = the **interior** color of shapes (bars, boxes, areas)
- **`color`** = the **outline** (border) of shapes, or the color of points and lines

For a histogram, `fill` colors the bars and `color` draws their borders.
:::

```{r}
ggplot(diabetes, aes(x = glucose_mg_dl)) +
  geom_histogram(binwidth = 10, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of Plasma Glucose",
    x = "Glucose (mg/dL)",
    y = "Number of patients"
  ) +
  theme_minimal()
```

We added three new elements:

- **`fill = "steelblue"`** --- bar interior color (set outside `aes()` because it's a fixed value, not mapped to a variable)
- **`labs(...)`** --- title and axis labels
- **`theme_minimal()`** --- a clean, uncluttered theme

### Adding a clinical reference line

Clinical thresholds are often worth marking on a plot. `geom_vline()` draws a vertical line:

```{r}
ggplot(diabetes, aes(x = glucose_mg_dl)) +
  geom_histogram(binwidth = 10, fill = "steelblue", color = "white") +
  geom_vline(xintercept = 140, linetype = "dashed", color = "firebrick") +
  labs(
    title = "Distribution of Plasma Glucose",
    subtitle = "Dashed line at 140 mg/dL (diabetes screening threshold)",
    x = "Glucose (mg/dL)",
    y = "Number of patients"
  ) +
  theme_minimal()
```

::: {.callout-tip}
## Reference lines for clinical thresholds

`geom_vline(xintercept = ...)` for vertical lines and `geom_hline(yintercept = ...)` for horizontal lines are great for marking clinical cutoffs --- fasting glucose thresholds, BMI boundaries, blood pressure targets, etc.
:::

### Interpretation

The glucose distribution is right-skewed: most patients cluster between 80--140 mg/dL, with a long tail extending toward higher values. The dashed line at 140 mg/dL shows where the clinical threshold falls.


## Comparing Groups: Boxplot {#sec-boxplot}

When comparing a continuous variable across groups, boxplots are often the best starting point. They show the **median**, **quartiles**, and **spread** in one compact display.

### A basic boxplot

```{r}
ggplot(diabetes, aes(x = diabetes_5y, y = glucose_mg_dl)) +
  geom_boxplot()
```

Already useful --- we can see the `pos` group has a higher median glucose.

### Adding color by group

Map `fill` to the grouping variable inside `aes()` to color each group:

```{r}
ggplot(diabetes, aes(x = diabetes_5y, y = glucose_mg_dl, fill = diabetes_5y)) +
  geom_boxplot() +
  labs(
    title = "Glucose by 5-Year Diabetes Outcome",
    x = "Outcome at 5 years",
    y = "Glucose (mg/dL)",
    fill = "Outcome"
  ) +
  theme_minimal()
```

::: {.callout-note}
## `fill` inside vs outside `aes()`

- **Inside `aes(fill = diabetes_5y)`** --- color is **mapped to a variable**. Each group gets its own color automatically.
- **Outside `aes()`, e.g., `fill = "steelblue"`** --- a **fixed color** applied to everything.

This inside/outside distinction applies to all aesthetics: `fill`, `color`, `size`, `alpha`, etc.
:::

### Overlaying individual points with `geom_jitter()`

Boxplots summarize, but they hide individual data points. Adding `geom_jitter()` shows every patient:

```{r}
ggplot(diabetes, aes(x = diabetes_5y, y = glucose_mg_dl, fill = diabetes_5y)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.25, size = 1) +
  labs(
    title = "Glucose by 5-Year Diabetes Outcome",
    x = "Outcome at 5 years",
    y = "Glucose (mg/dL)",
    fill = "Outcome"
  ) +
  theme_minimal()
```

- **`geom_jitter(width = 0.15)`** --- adds points with slight horizontal randomness to reduce overlap
- **`alpha = 0.25`** --- makes points semi-transparent so dense areas are visible
- **`outlier.shape = NA`** --- hides the boxplot's built-in outlier dots (since jitter already shows all points)

### Handling missing values with `drop_na()` {#sec-drop-na}

Real data has missing values. When ggplot2 encounters `NA`s, it usually drops them with a warning message. You can handle this explicitly with `drop_na()`:

```{r}
#| eval: false
# Drop rows where insulin is missing before plotting
diabetes |>
  drop_na(insulin_microiu_ml) |>
  ggplot(aes(x = diabetes_5y, y = insulin_microiu_ml)) +
  geom_boxplot()
```

`drop_na(column1, column2)` removes rows where **any** of the specified columns are `NA`. If you call `drop_na()` without arguments, it removes rows with `NA` in **any** column.

::: {.callout-warning}
## `drop_na()` removes entire rows

When you drop rows with missing values, you're removing those patients from the analysis entirely. This is fine for plotting, but be mindful of how much data you're losing --- especially if missingness is substantial (e.g., insulin has ~49% missing in this dataset).
:::

### Interpretation

The `pos` group clearly shows higher glucose values than the `neg` group, with the median shifted upward. The jitter points reveal that both groups have a wide spread, with considerable overlap.


## Exploring Relationships: Scatter Plot

When you want to see how **two continuous variables** relate, a scatter plot is the tool.

### A basic scatter plot

```{r}
ggplot(diabetes, aes(x = bmi, y = glucose_mg_dl)) +
  geom_point()
```

Each dot is one patient. We can already see a general upward trend.

### Adding color for a third variable

Add `color = diabetes_5y` inside `aes()` to show a third variable:

```{r}
ggplot(diabetes, aes(x = bmi, y = glucose_mg_dl, color = diabetes_5y)) +
  geom_point(alpha = 0.6) +
  labs(
    title = "BMI vs Glucose",
    x = "Body mass index (kg/m²)",
    y = "Glucose (mg/dL)",
    color = "Outcome"
  ) +
  theme_minimal()
```

::: {.callout-note}
## `fill` vs `color` --- when to use which

- Use **`color`** for **points** and **lines** (things without an interior)
- Use **`fill`** for **bars**, **boxes**, and **areas** (things with an interior)

Some geoms use both: `geom_boxplot()` uses `fill` for the box interior and `color` for the border.
:::

### Adding a trend line

`geom_smooth()` adds a fitted trend line. By default, it uses a flexible smoother. Add `method = "lm"` for a straight line:

```{r}
ggplot(diabetes, aes(x = bmi, y = glucose_mg_dl, color = diabetes_5y)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm") +
  labs(
    title = "BMI vs Glucose with Trend Lines",
    x = "Body mass index (kg/m²)",
    y = "Glucose (mg/dL)",
    color = "Outcome"
  ) +
  theme_minimal()
```

::: {.callout-note}
## What does `method = "lm"` mean?

`"lm"` stands for **linear model** --- it draws a straight-line best fit. The shaded band around the line is a **confidence interval**, showing the uncertainty in the trend. We'll learn about linear models formally in a later chapter; for now, just think of it as "draw the best straight line through these points."

If you omit `method = "lm"`, ggplot2 uses a flexible curve (called LOESS) that follows the data more closely.
:::

### Interpretation

Both groups show a slight positive trend: higher BMI tends to associate with higher glucose. The `pos` group's line sits higher overall. Remember, this is descriptive --- we're not claiming BMI *causes* higher glucose.

::: {.callout-caution collapse="true" title="Python Comparison"}
Python's seaborn provides similar scatter + trend line functionality:

```python
import seaborn as sns

sns.lmplot(
    data=diabetes,
    x="bmi", y="glucose_mg_dl",
    hue="diabetes_5y",
    height=5,
    scatter_kws={"alpha": 0.5}
)
```

`lmplot()` combines scatter points and linear regression lines, similar to `geom_point()` + `geom_smooth(method = "lm")`.
:::


## Categorical Comparisons: Bar Chart

For comparing **counts** across categories, bar charts are the standard.

### Preparing counts with `count()`

First, compute the counts (using `count()` from @sec-summarise):

```{r}
bmi_counts <- diabetes |>
  count(diabetes_5y, bmi_class)
bmi_counts
```

### `geom_col()` for pre-computed values

Since we already have the `n` column, we use `geom_col()` (not `geom_bar()`):

```{r}
ggplot(bmi_counts, aes(x = bmi_class, y = n, fill = diabetes_5y)) +
  geom_col(position = "dodge") +
  labs(
    title = "BMI Class by Diabetes Outcome",
    x = "BMI class",
    y = "Number of patients",
    fill = "Outcome"
  ) +
  theme_minimal()
```

::: {.callout-note}
## `geom_col()` vs `geom_bar()`

- **`geom_col()`** --- you provide both `x` and `y`. The bar heights come from your data (pre-computed counts).
- **`geom_bar()`** --- you provide only `x`. ggplot2 counts the rows for you automatically.

When you've already computed counts with `count()` or `summarise()`, use `geom_col()`.
:::

### Side-by-side bars with `position = "dodge"`

The `position` argument controls how grouped bars are arranged:

::: {.callout-tip}
## Three bar position options

| Position | Effect | Best for |
|---|---|---|
| `"dodge"` | Side-by-side | Comparing absolute counts |
| `"stack"` (default) | Stacked on top | Showing totals with group composition |
| `"fill"` | Stacked to 100% | Comparing proportions |
:::

```{r}
# Proportional view — each bar reaches 100%
ggplot(bmi_counts, aes(x = bmi_class, y = n, fill = diabetes_5y)) +
  geom_col(position = "fill") +
  labs(
    title = "Proportion of Diabetes Outcome by BMI Class",
    x = "BMI class",
    y = "Proportion",
    fill = "Outcome"
  ) +
  theme_minimal()
```

### Interpretation

The Obesity category dominates both groups, but the proportion of `pos` outcomes increases with higher BMI class.

::: {.callout-caution collapse="true" title="Python Comparison"}
Python's seaborn provides `countplot()` for similar bar charts:

```python
import seaborn as sns

sns.countplot(data=diabetes, x="bmi_class", hue="diabetes_5y")
```

For pre-computed values, use `sns.barplot()` or `matplotlib`'s `bar()`.
:::


## Splitting into Panels: Faceting

Sometimes overlaying groups on one plot gets crowded. **Faceting** splits the plot into separate panels --- one per group.

### `facet_wrap()` --- one variable

```{r}
ggplot(diabetes, aes(x = glucose_mg_dl)) +
  geom_histogram(binwidth = 10, fill = "steelblue", color = "white") +
  facet_wrap(~ diabetes_5y) +
  labs(
    title = "Glucose Distribution by Diabetes Outcome",
    x = "Glucose (mg/dL)",
    y = "Number of patients"
  ) +
  theme_minimal()
```

The syntax `facet_wrap(~ variable)` means "create one panel per level of this variable." The `~` (tilde) is R's formula notation --- here it simply means "split by."

::: {.callout-tip}
## When to facet vs when to color

- **Facet** when groups have very different distributions or when overlap makes the plot hard to read
- **Color** when groups overlap but you want them on the same axes for direct comparison

Both are valid --- choose whichever communicates more clearly.
:::


## Customizing Your Plots

### Labels with `labs()`

`labs()` sets all text elements at once:

```{r}
#| eval: false
labs(
  title    = "Main title",
  subtitle = "Additional context",
  x        = "X-axis label",
  y        = "Y-axis label",
  fill     = "Legend title (for fill)",
  color    = "Legend title (for color)",
  caption  = "Source or note"
)
```

Always include units in axis labels (e.g., "Glucose (mg/dL)" not just "Glucose").

### Themes: built-in options

ggplot2 includes several themes that change the overall look:

| Theme | Style |
|---|---|
| `theme_minimal()` | Clean, no box frame |
| `theme_bw()` | White background with border |
| `theme_classic()` | Publication-style with axes only |
| `theme_light()` | Light gray background, fine grid |
| `theme_void()` | Nothing but the data (useful for maps) |

: Common ggplot2 themes {.striped}

::: {.callout-tip}
## Choosing a theme

- For presentations: `theme_minimal()` (clean, modern)
- For publications: `theme_classic()` or `theme_bw()` (formal, traditional)
- For dashboards: `theme_light()` (readable on screens)

Pick one and use it consistently throughout a report.
:::

### Manual colors

Control colors explicitly with `scale_fill_manual()` or `scale_color_manual()`:

```{r}
#| eval: false
# For fill aesthetic (bars, boxes)
scale_fill_manual(values = c("neg" = "forestgreen", "pos" = "firebrick"))

# For color aesthetic (points, lines)
scale_color_manual(values = c("neg" = "forestgreen", "pos" = "firebrick"))
```

::: {.callout-note}
## `scale_fill_manual()` vs `scale_color_manual()`

Use `scale_fill_manual()` when your aesthetic is `fill` (bars, boxes). Use `scale_color_manual()` when your aesthetic is `color` (points, lines). They work identically, just for different aesthetics.
:::

### Journal-style palettes with `ggsci` (optional)

The `ggsci` package provides color palettes inspired by scientific journals:

```{r}
library(ggsci)

ggplot(diabetes, aes(x = bmi, y = glucose_mg_dl, color = diabetes_5y)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm") +
  scale_color_lancet() +
  labs(
    title = "BMI vs Glucose (Lancet palette)",
    x = "Body mass index (kg/m²)",
    y = "Glucose (mg/dL)",
    color = "Outcome"
  ) +
  theme_minimal()
```

`scale_color_lancet()` applies colors inspired by The Lancet journal. Other options include `scale_color_nejm()` (NEJM), `scale_color_npg()` (Nature), and `scale_color_jama()` (JAMA).

::: {.callout-tip}
## When to use `ggsci`

`ggsci` palettes are designed for a small number of groups (2--8). They're a quick way to get publication-quality colors without choosing hex codes manually.
:::


## Putting It All Together

Let's combine data transformation and visualization in one pipeline --- the real-world pattern:

```{r}
diabetes |>
  filter(bmi > 0) |>
  mutate(bmi_group = if_else(bmi >= 30, "Obese", "Non-obese")) |>
  ggplot(aes(x = glucose_mg_dl, fill = bmi_group)) +
  geom_histogram(binwidth = 10, color = "white", alpha = 0.7, position = "identity") +
  labs(
    title = "Glucose Distribution by Obesity Status",
    x = "Glucose (mg/dL)",
    y = "Number of patients",
    fill = "BMI group"
  ) +
  theme_minimal()
```

Notice the seamless transition from `|>` (data manipulation) to `+` (ggplot2 layers). The data flows through `filter()` and `mutate()`, then into `ggplot()`.

::: {.callout-tip}
## The transform-then-plot pattern

In practice, you'll often pipe data through a few dplyr verbs before handing it to `ggplot()`. This keeps your plotting code focused on *visualization* while the dplyr steps handle *data preparation*.
:::


## Common Errors and Troubleshooting

| Error or Symptom | Cause | Fix |
|---|---|---|
| Empty plot (gray box, no data) | Forgot `geom_*()` layer | Add `+ geom_histogram()` or similar |
| `object 'colname' not found` | Column name typo in `aes()` | Check `names(data)` |
| `stat_bin() requires a continuous x` | Used a factor/character on x of histogram | Check `class(column)`, convert if needed |
| `\|>` used instead of `+` | Mixed pipe and ggplot operators | Use `+` between ggplot layers |
| Legend shows unexpected groups | NA values in grouping variable | Use `drop_na()` before plotting |
| Bars overlap instead of side-by-side | Missing `position = "dodge"` | Add `position = "dodge"` to `geom_col()` |

: Common ggplot2 errors and their fixes {.striped}


## Summary

Here's the visualization journey we took with the diabetes data:

| Step | What we visualized | ggplot2 tool |
|---|---|---|
| Distribution | Glucose histogram | `geom_histogram()` |
| Reference | Clinical threshold line | `geom_vline()` |
| Group comparison | Glucose by outcome (boxplot) | `geom_boxplot()`, `geom_jitter()` |
| Relationship | BMI vs glucose (scatter) | `geom_point()`, `geom_smooth()` |
| Categories | BMI class by outcome (bars) | `geom_col()`, `position` |
| Panels | Glucose faceted by outcome | `facet_wrap()` |
| Style | Labels, themes, palettes | `labs()`, `theme_*()`, `ggsci` |

: Chapter 5 journey --- from raw data to polished plots {.striped}

In the next chapter, we'll move from visual summaries to formal **descriptive statistics** --- computing and presenting the numbers behind what we've been seeing in these plots.


## Exercises {.unnumbered}

1. **Histogram with threshold.** Create a histogram of `dbp_mm_hg` (diastolic blood pressure) with bin width 5. Add a vertical dashed line at 80 mmHg (the threshold for elevated blood pressure). Include a title and axis labels.

2. **Group comparison.** Create a boxplot of `bmi` by `diabetes_5y`, with jitter points overlaid. Write 2--3 sentences interpreting what you observe about the two groups.

3. **Scatter with trend.** Plot `age` (x-axis) vs `glucose_mg_dl` (y-axis), colored by `diabetes_5y`. Add linear trend lines. Does the relationship between age and glucose appear different for the two outcome groups?

4. **Faceted bar chart.** Compute counts of `bmi_class` by `diabetes_5y`, then create a dodged bar chart. Now recreate it using `position = "fill"` to show proportions. Which version better highlights the relationship between BMI class and diabetes outcome?
