# Descriptive Statistics

In this chapter, we focus on practical implementation of descriptive statistics in R. The emphasis is on reusable coding patterns for clinical datasets, rather than detailed statistical theory.

By the end of this chapter, you should be able to quickly generate:

- continuous-variable summaries,
- categorical n (%) summaries,
- grouped summaries by outcome,
- missingness reports,
- and a publication-ready baseline table.

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(janitor)
library(gtsummary)
```

::: {.callout-note}
## Running example: diabetes registry summary

We will continue using `data/diabetes.csv` and answer one practical question:

**How can we summarize this cohort in a clear, reproducible, report-ready format?**
:::


## Data Setup for This Chapter

```{r}
diabetes_raw <- read_csv("../../data/diabetes.csv", show_col_types = FALSE)

diabetes <- diabetes_raw |>
  clean_names() |>
  mutate(
    diabetes_5y = fct_relevel(diabetes_5y, "neg", "pos"),
    bmi_class = case_when(
      bmi < 18.5 ~ "Underweight",
      bmi < 25 ~ "Normal",
      bmi < 30 ~ "Overweight",
      TRUE ~ "Obesity"
    ) |> factor(levels = c("Underweight", "Normal", "Overweight", "Obesity"))
  )

glimpse(diabetes)
```

### Syntax notes (data setup)

- **`clean_names()`** standardizes variable names for easier coding.
- **`fct_relevel(diabetes_5y, "neg", "pos")`** sets category order explicitly.
- **`case_when(...)`** creates a new categorical variable using rule-based conditions.
- **`factor(levels = ...)`** fixes display/reporting order for categories.


## 1) Continuous Variables: Fast Summary in R

For continuous variables, the most common descriptive statistics are:

- mean and standard deviation (SD),
- median and interquartile range (IQR),
- minimum and maximum.

### Single variable example

```{r}
mean(diabetes$glucose_mg_dl, na.rm = TRUE)
sd(diabetes$glucose_mg_dl, na.rm = TRUE)
median(diabetes$glucose_mg_dl, na.rm = TRUE)
IQR(diabetes$glucose_mg_dl, na.rm = TRUE)
```

### Multiple variables with one reusable pattern

```{r}
diabetes |>
  summarise(
    across(
      .cols = c(age, bmi, glucose_mg_dl, dbp_mm_hg),
      .fns = list(
        n = ~sum(!is.na(.x)),
        mean = ~mean(.x, na.rm = TRUE),
        sd = ~sd(.x, na.rm = TRUE),
        median = ~median(.x, na.rm = TRUE),
        iqr = ~IQR(.x, na.rm = TRUE),
        min = ~min(.x, na.rm = TRUE),
        max = ~max(.x, na.rm = TRUE)
      ),
      .names = "{.col}_{.fn}"
    )
  )
```

### Syntax notes (new pattern)

- **`across()`** applies the same operations to multiple columns.
- **`.fns = list(...)`** defines several summary functions at once.
- **`.names = "{.col}_{.fn}"`** controls output column names.
- **`na.rm = TRUE`** excludes missing values from computations.

Notable options to explore:

- Use **`across(where(is.numeric), ...)`** when you want to summarize all numeric columns automatically.
- Use **`round()`** inside summary functions (or after summarization) when preparing reader-facing tables.

::: {.callout-tip}
## Practical reporting note

For skewed biomedical variables (e.g., insulin), median and IQR are often more robust than mean and SD.
:::


## 2) Categorical Variables: n and %

For categorical variables, report frequencies and proportions.

```{r}
diabetes |>
  count(diabetes_5y)
```

```{r}
diabetes |>
  count(diabetes_5y) |>
  mutate(prop_pct = 100 * n / sum(n))
```

Now a grouped categorical example:

```{r}
diabetes |>
  count(diabetes_5y, bmi_class) |>
  group_by(diabetes_5y) |>
  mutate(prop_pct = 100 * n / sum(n)) |>
  ungroup()
```

### Syntax notes

- **`count(a, b)`** is a compact form of `group_by(a, b) |> summarise(n = n())`.
- **`mutate(prop_pct = ...)`** computes percentages explicitly.
- Always ensure the denominator is clear (`sum(n)` within each group).

Notable options to explore:

- **`count(..., sort = TRUE)`** sorts counts from highest to lowest.
- **`count(..., name = "freq")`** customizes the count column name.


## 3) Grouped Descriptive Summaries

A common clinical workflow is to summarize key variables by outcome group.

```{r}
diabetes |>
  group_by(diabetes_5y) |>
  summarise(
    n = n(),
    mean_age = mean(age, na.rm = TRUE),
    mean_bmi = mean(bmi, na.rm = TRUE),
    mean_glucose = mean(glucose_mg_dl, na.rm = TRUE),
    median_glucose = median(glucose_mg_dl, na.rm = TRUE),
    .groups = "drop"
  )
```

### Syntax notes

- **`group_by(diabetes_5y)`** performs summaries within each outcome group.
- **`.groups = "drop"`** returns an ungrouped tibble after summarization, which avoids accidental carry-over of grouping in later code.

This produces a concise group-level profile and is often the first summary table used in reporting.


## 4) Missing Data Summary (Implementation Pattern)

Missing data should be explicitly reported.

```{r}
missingness_tbl <- tibble(variable = names(diabetes)) |>
  mutate(
    n_missing = map_int(variable, ~sum(is.na(diabetes[[.x]]))),
    pct_missing = 100 * n_missing / nrow(diabetes)
  ) |>
  arrange(desc(pct_missing))

missingness_tbl
```

```{r}
missingness_tbl |>
  slice_head(n = 5)
```

Notable option to explore:

- For numeric-only missing checks, you can also use a compact pattern such as `summarise(across(where(is.numeric), ~sum(is.na(.x))))`.

This pattern is simple, transparent, and useful in methods/results sections.


## 5) Publication-Ready Baseline Table with `gtsummary`

At this point, we have the core summaries. Now we format them for publication-style presentation.

```{r}
diabetes |>
  select(diabetes_5y, age, bmi, glucose_mg_dl, dbp_mm_hg, bmi_class) |>
  tbl_summary(
    by = diabetes_5y,
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = all_continuous() ~ 1,
    missing = "ifany",
    label = list(
      age ~ "Age (years)",
      bmi ~ "Body mass index (kg/m^2)",
      glucose_mg_dl ~ "Glucose (mg/dL)",
      dbp_mm_hg ~ "Diastolic blood pressure (mmHg)",
      bmi_class ~ "BMI class"
    )
  ) |>
  add_overall(last = TRUE) |>
  bold_labels()
```

### Syntax notes

- **`tbl_summary(by = ...)`** creates a stratified baseline table.
- **`statistic = list(...)`** controls display format.
- **`add_overall()`** adds the overall cohort column.
- **`bold_labels()`** improves readability.

Notable options to explore:

- For skewed variables, you may prefer `all_continuous() ~ "{median} ({p25}, {p75})"`.
- `missing = "no"` hides missing rows; `missing = "always"` always shows them.
- In inferential chapters, `add_p()` can be added when you are ready to report statistical tests.

::: {.callout-note}
## Optional extension

If you want to export this table to file formats (HTML/Word), you can convert it with `as_gt()` and save with `gt` workflows.
:::


## End-to-End Mini Workflow

A compact implementation pattern for descriptive reporting:

```{r}
diabetes |>
  filter(glucose_mg_dl > 0, bmi > 0) |>
  group_by(diabetes_5y) |>
  summarise(
    n = n(),
    mean_glucose = mean(glucose_mg_dl, na.rm = TRUE),
    mean_bmi = mean(bmi, na.rm = TRUE),
    .groups = "drop"
  )
```

Then convert to a publication-ready table with `tbl_summary()` when needed.


## Summary

| Task | Recommended R pattern |
|---|---|
| Continuous summaries | `summarise(across(...))` |
| Categorical summaries | `count(...)` + percentage with `mutate()` |
| Grouped summaries | `group_by(...)` + `summarise(...)` |
| Missingness | `is.na()` + variable-wise tally |
| Publication-ready baseline table | `gtsummary::tbl_summary()` |

In the next chapter, we move from descriptive summaries to **inferential comparison for two groups**.


## Exercises {.unnumbered}

1. **Continuous summary practice.** Using `summarise(across())`, compute mean, SD, median, and IQR for `age`, `bmi`, and `glucose_mg_dl`.

2. **Categorical profile.** Report n (%) of `bmi_class` overall and stratified by `diabetes_5y`.

3. **Missingness table.** Create a table with `variable`, `n_missing`, and `pct_missing`, sorted descending by `pct_missing`.

4. **Baseline table challenge.** Build a `gtsummary` table stratified by `diabetes_5y`, include at least four variables, and use custom labels with one decimal for continuous summaries.

Solutions will be provided in Appendix C.
