# Descriptive Statistics

Before running any statistical test, you need to **describe** your data. How old are the patients? What is the distribution of glucose? How much data is missing? Is the cohort balanced across groups?

In this chapter, you'll learn to compute and present descriptive summaries --- the kind you'd include in a methods section, a departmental presentation, or a publication's "Table 1."

```{r}
#| message: false
library(tidyverse)
library(janitor)
library(gtsummary)
```

::: {.callout-note}
## Our running example: Summarizing the diabetes cohort

Imagine you're preparing a summary for a departmental meeting. You need to describe the 768 patients in the diabetes registry:

- What are the typical values for age, BMI, and glucose?
- How many patients developed diabetes within 5 years?
- How much data is missing?
- Can you produce a single, publication-ready table?

We'll answer each of these questions step by step.
:::


## Data Preparation

We load and prepare the diabetes data using the same tools from Chapters 4 and 5:

```{r}
diabetes_raw <- read_csv("../../data/diabetes.csv", show_col_types = FALSE)

diabetes <- diabetes_raw |>
  clean_names() |>
  mutate(
    diabetes_5y = fct_relevel(diabetes_5y, "neg", "pos"),
    bmi_class = case_when(
      bmi < 18.5 ~ "Underweight",
      bmi < 25   ~ "Normal",
      bmi < 30   ~ "Overweight",
      .default   = "Obesity"
    ) |> factor(levels = c("Underweight", "Normal", "Overweight", "Obesity"))
  )
```


## Choosing the Right Summary Statistics

Before computing anything, it helps to know **which statistics suit which data**. The choice depends on the variable type and its distribution:

| Variable Type | Symmetric Distribution | Skewed Distribution |
|---|---|---|
| **Continuous** (e.g., age, BMI) | Mean and SD | Median and IQR |
| **Categorical** (e.g., sex, outcome) | n and % | n and % |

: Choosing the right summary statistics {.striped}

For continuous variables, here's what each statistic measures:

| Statistic | What it measures | R function |
|---|---|---|
| Mean | Average (center of gravity) | `mean()` |
| SD (standard deviation) | Spread around the mean | `sd()` |
| Median | Middle value (50th percentile) | `median()` |
| IQR (interquartile range) | Spread of the middle 50% | `IQR()` |
| Min / Max | Extremes | `min()`, `max()` |

: Common summary statistics and their R functions {.striped}

::: {.callout-tip}
## Mean + SD vs Median + IQR

- Use **Mean + SD** when the data is roughly symmetric (bell-shaped)
- Use **Median + IQR** when the data is skewed (long tail on one side) or has outliers

For clinical data, many variables (like insulin, length of stay, costs) are right-skewed, making median + IQR the safer default. When in doubt, check with a histogram (@sec-ggplot2).
:::


## Summarizing Continuous Variables

### Single variable summary

Let's start with glucose. You can call summary functions directly, just like in @sec-vectors:

```{r}
mean(diabetes$glucose_mg_dl, na.rm = TRUE)
```

```{r}
sd(diabetes$glucose_mg_dl, na.rm = TRUE)
```

```{r}
median(diabetes$glucose_mg_dl, na.rm = TRUE)
```

```{r}
IQR(diabetes$glucose_mg_dl, na.rm = TRUE)
```

Recall from @sec-na that `na.rm = TRUE` tells R to skip missing values when computing the statistic.

### Summarizing multiple variables

To summarize several variables at once, use `summarise()` with one line per statistic:

```{r}
diabetes |>
  summarise(
    mean_age      = mean(age, na.rm = TRUE),
    sd_age        = sd(age, na.rm = TRUE),
    mean_bmi      = mean(bmi, na.rm = TRUE),
    sd_bmi        = sd(bmi, na.rm = TRUE),
    mean_glucose  = mean(glucose_mg_dl, na.rm = TRUE),
    sd_glucose    = sd(glucose_mg_dl, na.rm = TRUE)
  )
```

This is verbose but completely transparent --- you can see exactly what's being computed for each variable.

::: {.callout-tip collapse="true"}
## Shortcut: `across()` for many columns

When you have many columns to summarize, the `across()` function can save typing. This uses more advanced syntax that we won't cover in detail --- just know it exists:

```{r}
diabetes |>
  summarise(
    across(
      c(age, bmi, glucose_mg_dl),
      list(mean = \(x) mean(x, na.rm = TRUE),
           sd   = \(x) sd(x, na.rm = TRUE))
    )
  )
```

The column-by-column approach above is perfectly fine for most analyses. Use `across()` when you're comfortable with it.
:::

::: {.callout-caution collapse="true" title="Python Comparison"}
Python's pandas provides `describe()` for quick summaries:

```python
diabetes[["age", "bmi", "glucose_mg_dl"]].describe()
```

This returns count, mean, std, min, percentiles, and max --- similar to combining `mean()`, `sd()`, `median()`, `min()`, `max()` in R.
:::


## Summarizing Categorical Variables

### Frequency table with `count()`

For categorical variables, the key statistics are **n** (count) and **%** (proportion). Use `count()` from @sec-summarise:

```{r}
diabetes |> count(diabetes_5y)
```

### Adding percentages

Add a percentage column with `mutate()`:

```{r}
diabetes |>
  count(diabetes_5y) |>
  mutate(pct = round(100 * n / sum(n), 1))
```

About 35% of patients in this cohort developed diabetes within 5 years.

### Stratified frequency table

To see how one categorical variable breaks down **within** another group, use `count()` with two variables and then compute within-group percentages:

```{r}
diabetes |>
  count(diabetes_5y, bmi_class) |>
  group_by(diabetes_5y) |>
  mutate(pct = round(100 * n / sum(n), 1)) |>
  ungroup()
```

::: {.callout-warning}
## Watch your denominator

When computing percentages, be clear about **what the denominator is**:

- `sum(n)` after `group_by(diabetes_5y)` → percentages **within each outcome group** (columns sum to 100%)
- `sum(n)` without grouping → percentages **of the total cohort** (all rows sum to 100%)

Getting the denominator wrong is one of the most common mistakes in clinical reporting.
:::


## Grouped Summaries by Outcome

A key step in any clinical analysis: summarize continuous variables **by group** to see how the groups compare.

### Continuous variables by group

```{r}
diabetes |>
  group_by(diabetes_5y) |>
  summarise(
    n            = n(),
    mean_age     = round(mean(age, na.rm = TRUE), 1),
    sd_age       = round(sd(age, na.rm = TRUE), 1),
    mean_glucose = round(mean(glucose_mg_dl, na.rm = TRUE), 1),
    sd_glucose   = round(sd(glucose_mg_dl, na.rm = TRUE), 1),
    mean_bmi     = round(mean(bmi, na.rm = TRUE), 1),
    sd_bmi       = round(sd(bmi, na.rm = TRUE), 1),
    .groups = "drop"
  )
```

### Interpreting group differences

The `pos` group is older on average (37 vs 31 years), has higher mean glucose (142 vs 110 mg/dL), and higher mean BMI (35 vs 30 kg/m²). These differences are descriptive --- we haven't tested whether they're statistically significant. That comes in the next chapters.

::: {.callout-caution collapse="true" title="Python Comparison"}
Python's `groupby` + `agg` achieves the same result:

```python
(
    diabetes
    .groupby("diabetes_5y")
    .agg(
        n=("age", "size"),
        mean_age=("age", "mean"),
        mean_glucose=("glucose_mg_dl", "mean"),
        mean_bmi=("bmi", "mean")
    )
    .round(1)
)
```
:::


## Assessing Missing Data

### Why report missingness?

Missing data is common in clinical datasets and can bias your results. Reporting how much data is missing --- and in which variables --- is essential for transparency. Journals and reviewers expect it.

### Counting missing values per variable

The simplest approach: use `summarise()` with `is.na()` for each column you care about:

```{r}
diabetes |>
  summarise(
    glucose_missing     = sum(is.na(glucose_mg_dl)),
    dbp_missing         = sum(is.na(dbp_mm_hg)),
    triceps_missing     = sum(is.na(triceps_mm)),
    insulin_missing     = sum(is.na(insulin_microiu_ml)),
    bmi_missing         = sum(is.na(bmi))
  )
```

`sum(is.na(x))` counts the `TRUE`s --- recall from @sec-logical-ops that `TRUE` counts as 1. So `sum(is.na(x))` is literally "how many values are NA?"

### A missingness summary table

For a cleaner report, build a tibble with variable names, counts, and percentages:

```{r}
miss_table <- tibble(
  variable    = c("glucose_mg_dl", "dbp_mm_hg", "triceps_mm",
                  "insulin_microiu_ml", "bmi"),
  n_missing   = c(
    sum(is.na(diabetes$glucose_mg_dl)),
    sum(is.na(diabetes$dbp_mm_hg)),
    sum(is.na(diabetes$triceps_mm)),
    sum(is.na(diabetes$insulin_microiu_ml)),
    sum(is.na(diabetes$bmi))
  ),
  pct_missing = round(100 * n_missing / nrow(diabetes), 1)
) |>
  arrange(desc(pct_missing))

miss_table
```

::: {.callout-tip collapse="true"}
## Shortcut: using `across()` for missingness

For many columns, you can compute missingness more compactly:

```{r}
diabetes |>
  summarise(across(everything(), \(x) sum(is.na(x))))
```

This applies `sum(is.na())` to every column at once. Handy when you have dozens of variables.
:::

### Interpretation

Insulin has ~49% missing and triceps has ~30% missing --- that's substantial. When analyzing these variables, you'll need to consider whether the missingness is random or systematic. At a minimum, report it clearly.


## Publication-Ready Tables with `gtsummary` {#sec-gtsummary}

So far, we've built summaries manually with `summarise()`. That works well for understanding the data, but for a **publication-quality "Table 1"**, the `gtsummary` package automates the formatting.

::: {.callout-note}
## One function, one Table 1

`tbl_summary()` takes a data frame and produces a formatted summary table --- the kind you'd include in a journal article's "Baseline Characteristics" table. It handles continuous and categorical variables automatically.
:::

### Step 1: The simplest table

Start with just the defaults --- no customization at all:

```{r}
diabetes |>
  select(age, bmi, glucose_mg_dl, diabetes_5y) |>
  tbl_summary()
```

With one function call, `gtsummary` detected variable types, computed appropriate statistics (median + IQR for continuous, n + % for categorical), and formatted everything.

### Step 2: Stratify by outcome

Add `by = diabetes_5y` to split the table by group:

```{r}
diabetes |>
  select(diabetes_5y, age, bmi, glucose_mg_dl, bmi_class) |>
  tbl_summary(by = diabetes_5y)
```

Now each column is a group, and you can compare at a glance.

### Step 3: Customize the statistics

By default, `gtsummary` reports median (IQR) for continuous variables. To switch to mean (SD), use the `statistic` argument:

```{r}
diabetes |>
  select(diabetes_5y, age, bmi, glucose_mg_dl, bmi_class) |>
  tbl_summary(
    by = diabetes_5y,
    statistic = list(
      all_continuous()  ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = all_continuous() ~ 1
  )
```

Let's break down the new syntax:

- **`statistic = list(...)`** --- controls what statistics are displayed. Each entry maps a set of variables to a template string.
- **Template strings** use `{placeholder}` syntax:

| Placeholder | Meaning |
|---|---|
| `{mean}` | Mean |
| `{sd}` | Standard deviation |
| `{median}` | Median |
| `{p25}`, `{p75}` | 25th and 75th percentiles |
| `{n}` | Count |
| `{p}` | Percentage |

: `gtsummary` template placeholders {.striped}

- **`all_continuous()`** and **`all_categorical()`** are **selector helpers** --- they select all continuous or all categorical variables, similar to how `starts_with()` works in `select()`. Think of them as shortcuts for "apply this format to all variables of this type."
- **`digits = all_continuous() ~ 1`** --- round continuous statistics to 1 decimal place.

### Step 4: Add labels and polish

Add readable labels, an overall column, and bold formatting:

```{r}
diabetes |>
  select(diabetes_5y, age, bmi, glucose_mg_dl, dbp_mm_hg, bmi_class) |>
  tbl_summary(
    by = diabetes_5y,
    statistic = list(
      all_continuous()  ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = all_continuous() ~ 1,
    missing = "ifany",
    label = list(
      age           ~ "Age (years)",
      bmi           ~ "Body mass index (kg/m²)",
      glucose_mg_dl ~ "Glucose (mg/dL)",
      dbp_mm_hg     ~ "Diastolic BP (mmHg)",
      bmi_class     ~ "BMI class"
    )
  ) |>
  add_overall(last = TRUE) |>
  bold_labels()
```

New pieces:

- **`label = list(variable ~ "Label")`** --- gives each variable a human-readable name
- **`missing = "ifany"`** --- shows a "Missing" row only when there are actually missing values. Other options: `"no"` (hide missing rows) or `"always"` (always show)
- **`add_overall(last = TRUE)`** --- adds an "Overall" column at the end
- **`bold_labels()`** --- makes variable names bold for readability

This is a complete Table 1 --- ready for a report, slide deck, or journal submission.

::: {.callout-note}
## Exporting your table

To save the table as an HTML or Word file:

```{r}
#| eval: false
library(gt)

diabetes |>
  select(diabetes_5y, age, bmi, glucose_mg_dl) |>
  tbl_summary(by = diabetes_5y) |>
  as_gt() |>
  gtsave("table1.html")     # or "table1.docx" for Word
```

The `as_gt()` function converts the `gtsummary` table to a `gt` object, which can be saved in various formats.
:::

::: {.callout-note}
## What about p-values?

You may notice that Table 1 often includes a "p-value" column comparing groups. In `gtsummary`, you add this with `add_p()`. We'll cover this in the inferential statistics chapters --- for now, focus on describing the data clearly.
:::

::: {.callout-caution collapse="true" title="Python Comparison"}
Python's `tableone` package provides similar functionality:

```python
from tableone import TableOne

table1 = TableOne(
    data=diabetes,
    columns=["age", "bmi", "glucose_mg_dl", "bmi_class"],
    groupby="diabetes_5y",
    pval=False
)
print(table1.tabulate(tablefmt="github"))
```

R's `gtsummary` is generally more polished for publication-ready output, with built-in formatting and export options.
:::


## Common Errors and Troubleshooting

| Error or Symptom | Cause | Fix |
|---|---|---|
| `NaN` in mean or SD | All values are `NA` in that subset | Check your `filter()` --- are you accidentally removing all data? |
| Wrong percentages (don't sum to 100%) | Wrong `group_by()` before `mutate(pct = ...)` | Check your grouping --- denominator must match your intended total |
| `gtsummary` shows "Unknown" category | `NA` values in a categorical variable | Use `mutate(var = fct_na_value_to_level(var))` or `missing = "no"` |
| `na.rm` forgotten | `NA` result from `mean()`, `sd()`, etc. | Add `na.rm = TRUE` to every summary function |
| Very wide output from `summarise()` | Too many columns in one `summarise()` call | Break into separate summaries or use `gtsummary` instead |

: Common descriptive statistics errors and their fixes {.striped}


## Summary

Here's the journey we took to describe the diabetes cohort:

| Step | What we did | R tool |
|---|---|---|
| Choose statistics | Matched measure to data type | Mean + SD vs Median + IQR |
| Single variable | Computed glucose statistics | `mean()`, `sd()`, `median()`, `IQR()` |
| Multiple variables | Summarized age, BMI, glucose | `summarise()` column-by-column |
| Categorical counts | Frequency of outcome and BMI class | `count()`, `mutate(pct = ...)` |
| Grouped summaries | Compared pos vs neg groups | `group_by()` + `summarise()` |
| Missingness | Identified insulin as 49% missing | `sum(is.na())` per variable |
| Table 1 | Publication-ready baseline table | `gtsummary::tbl_summary()` |

: Chapter 6 journey --- from raw numbers to a publication-ready Table 1 {.striped}

These descriptive statistics are the foundation for any clinical analysis. In the next chapters, we'll move to **inferential statistics** --- testing whether the differences we've observed between groups are statistically significant.


## Exercises {.unnumbered}

1. **Vital signs summary.** Using the diabetes dataset, compute the mean, SD, median, and IQR of `age`, `bmi`, and `dbp_mm_hg` using explicit `summarise()` (one line per statistic). Round to 1 decimal.

2. **Categorical profile.** Create a frequency table of `bmi_class` with percentages. Then create a stratified version by `diabetes_5y` showing within-group percentages. Which BMI class is most common in the `pos` group?

3. **Missingness report.** Build a missingness summary table for all numeric columns in the diabetes dataset. Sort by percent missing (highest first). Which variable has the most missing data, and what percentage is missing?

4. **Table 1 challenge.** Build a `gtsummary` table stratified by `diabetes_5y` that includes `age`, `bmi`, `glucose_mg_dl`, `dbp_mm_hg`, and `bmi_class`. Use mean (SD) for continuous variables, custom labels with units, and add an overall column. Export it to HTML if you'd like to practice with `as_gt()`.
